### 2012-12-08 trunk/paper/paper.tex
Started paper. Well, outline.

\title{RADseq Works in Primates, Dammit.}
\author{
        Christina M. Bergey
			\and
			Andrew S. Burrell
			\and
			Luca S.J. Pozzi
			\and
			Todd, I suppose
	}
\date{\today}

\documentclass[12pt]{article}

\begin{document}
\maketitle

\begin{abstract}
\ldots Blah, blah, blah, RADseq, blah, blah, Cercopithecoidea. \ldots
\end{abstract}

\section{Introduction}
\begin{itemize}
	\item Next-gen sequencing revolution promises gains in primatology
	\item Still expensive
	\item Many genomes, but still tough doing genomics on non-model organisms
	\item What is RADseq?
	\item PRESENT STUDY
	\begin{itemize}
		\item We did RADseq on 6 Cercopithecoids
		\item Assessed how well it worked
		\item Did simple phylogeny?
	\end{itemize}
\end{itemize}

\section{Methods}

\paragraph{Library Preparation and Sequencing}
\begin{itemize}
	\item 6 animals, table with sources and other info
	\item Etter et al. 2011, though with modifications
	\item How we picked enzyme, PspXI
	\item Adapter sequences. Barcodes with at least 3 mismatches. PE adapter
	\item What did it look like on BioA? Size? Concentration?
	\item Sequenced on Illumina MiSeq. 150PE
	\item NYU Langone Medical Center's Genome Technology Center (name right?)
	\item How much actually loaded?
	\item 30\% spike in with PhiX control DNA to control for low diversity library
\end{itemize}

\paragraph{Analysis Pipeline - Mapping to Reference Genomes}
\begin{itemize}
	\item Demultiplex. Must have barcode and restriction site intact.
	\item Analyze reads with FastQC
	\begin{itemize}
		\item Total sequence bp
		\item Maximum possible sequence depth
		\item Other stats that FastQC gives you
	\end{itemize}
	\item Aligned to rhesus genome using BWA aln
	\begin{itemize}
		\item default parameters
	\end{itemize}
	\item Combine paired-end reads with BWA sampe
	\item Convert to BAM, sort and index with samtools
	\item Analyze mapped reads with samtools utilities flagstat and idxstats and bamtools utility 
	\item Post-alignment filtering steps
	\begin{itemize}
		\item Fix mate pair info with Picard
		\item Filter for mapped and paired.
		\item Remove dups with Picard
		\item Add read group info with Picard
		\item Remove reads with low mapping quality with bamtools
	\end{itemize}
\end{itemize}

\paragraph{Analysis Pipeline - Variant Calling}
\begin{itemize}
	\item Local realignment with GATK
	\item Fix paired end data with Picard
	\item Call SNPs with samtools
	\item Summarize SNP stats with vcf-stats
	%\item Call consensus sequence with samtools
\end{itemize}

\paragraph{Analysis Pipeline - Analysis of Degree of Overlap}
\begin{itemize}
	\item Calculate coverage of restriction site-associated regions
	\begin{itemize}
		\item Info on targeted intervals
		\begin{itemize}
			\item Total number possible targets in rhesus genome (compare to human too?)
			\item Total possible target BP
		\end{itemize}
		\item How many targets did we hit?
		\begin{itemize}
			\item BEDtools multiBamCoverage for this job
			\item Number and percentage of targets with coverage $\ge 1$
			\item Number and percentage of targets with coverage $\ge N$
		\end{itemize}
	\end{itemize}
	\item Count orthologous SNPs shared between individuals
	\begin{itemize}
		\item VCFtools vcf-compare for this job
	\end{itemize}
\end{itemize}

\paragraph{Analysis Pipeline - Inferring Phylogeny}
\begin{itemize}
	\item Using Stacks? 
	\item Using method like cichlid people?
	\item Using method like Rubin et al %http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0033394
\end{itemize}

\section{Results}
\begin{itemize}
	\item Table: 
	\begin{itemize}
		\item Number of reads per animal
		\item Number that passed filtration
		\item Number of loci hit
		\item Number of loci hit with coverage $\ge N$
		\item Number of SNPs
	\end{itemize}
	\item SNP Venn diagram?
	\item Table of overlapping region, orthologous SNP counts
	\item Phylogenetic tree
\end{itemize}

\section{Conclusions}
Conclusions

\section{Acknowledgements}
Acknowledgements

\end{document}

### 2012-12-08 trunk/paper/paper.tex
Added to paper outline a bit.

@@
 	\item What is RADseq?
+	\item Previous RADseq studies
+	\item Why would it be good for primates
 	\item PRESENT STUDY
@@
 		\item Assessed how well it worked
-		\item Did simple phylogeny?
+		\item Show it has promise for primates
 	\end{itemize}
@@
 	\item Etter et al. 2011, though with modifications
-	\item How we picked enzyme, PspXI
+	\item How we picked enzyme, PspXI. UCSC oligoMatch
 	\item Adapter sequences. Barcodes with at least 3 mismatches. PE adapter
@@
 	\item Demultiplex. Must have barcode and restriction site intact.
-	\item Analyze reads with FastQC
+	\item Analyze reads with FastQC (No need to say this)
 	\begin{itemize}
@@
 	\end{itemize}
-	\item Combine paired-end reads with BWA sampe
-	\item Convert to BAM, sort and index with samtools
+	\item Combine paired-end reads with BWA sampe (No need to say this)
+	\item Convert to BAM, sort and index with samtools (No need to say this)
 	\item Analyze mapped reads with samtools utilities flagstat and idxstats and bamtools utility 
@@
 		\item Remove dups with Picard
-		\item Add read group info with Picard
+		\item Add read group info with Picard (No need to say this)
 		\item Remove reads with low mapping quality with bamtools
@@
 	\item Local realignment with GATK
-	\item Fix paired end data with Picard
-	\item Call SNPs with samtools
-	\item Summarize SNP stats with vcf-stats
+	\item Fix paired end data with Picard (No need to say this)
+	\item Call SNPs with samtools. Min coverage = 3, max = 100
+	\item Summarize SNP stats with vcf-stats (No need to say this)
 	%\item Call consensus sequence with samtools
@@
 
-\paragraph{Analysis Pipeline - Inferring Phylogeny}
-\begin{itemize}
-	\item Using Stacks? 
-	\item Using method like cichlid people?
-	\item Using method like Rubin et al %http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0033394
-\end{itemize}
+%\paragraph{Analysis Pipeline - Inferring Phylogeny}
+%\begin{itemize}
+%	\item Using Stacks? 
+%	\item Using method like cichlid people?
+%	\item Using method like Rubin et al %http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0033394
+%\end{itemize}
 
@@
 	\item Table of overlapping region, orthologous SNP counts
-	\item Phylogenetic tree
+	%\item Phylogenetic tree
 \end{itemize}
 
-\section{Conclusions}
-Conclusions
+\section{Discussion}
+\begin{itemize}
+	\item RADseq is viable tool for researcher interested in primate phylogenetics, pop. gen.
+	\item Enzyme choice allows control over coverage, number of individuals, number of loci.
+	\item Potential problems with RADseq method
+	\item Promise for primatology
+\end{itemize}
 

### 2012-12-08 trunk/paper/paper.tex
Added library prep and sequencing to paper

@@
 \documentclass[12pt]{article}
+
+\usepackage{color}
 
@@
 \paragraph{Library Preparation and Sequencing}
-\begin{itemize}
-	\item 6 animals, table with sources and other info
-	\item Etter et al. 2011, though with modifications
-	\item How we picked enzyme, PspXI. UCSC oligoMatch
-	\item Adapter sequences. Barcodes with at least 3 mismatches. PE adapter
-	\item What did it look like on BioA? Size? Concentration?
-	\item Sequenced on Illumina MiSeq. 150PE
-	\item NYU Langone Medical Center's Genome Technology Center (name right?)
-	\item How much actually loaded?
-	\item 30\% spike in with PhiX control DNA to control for low diversity library
-\end{itemize}
+
+	Genomic DNA from 6 animals was digested with \emph{PspXI} (New England Biolabs) and used to create a multiplexed RAD tag library.
+Our library preparation method followed that of Etter et al, 2011 with the following modifications: the P1 adapter top\textcolor{red}{(?)} oligonucleotide was modified to have an overhang corresponding to the cut site of \emph{PspXI}, and a longer P2 adapter suitable for paired end sequencing was used (P2\_top: 5'-\textcolor{red}{SEQUENCEHERE}-3'; P2\_bottom: 5'-\textcolor{red}{SEQUENCEHERE}-3').
+Individual-specific barcodes contained in the P1 adapter differed by at least \textcolor{red}{three} nucleotides.
+We chose \emph{PspXI} based on the results of \emph{in silico} digestion of the human, rhesus macaque, and baboon reference genomes using custom Perl scripts (\textcolor{red}{refs}).
+We sequenced the prepared library as one 150-cycle paired-end run of an Illumina MiSeq at the NYU Langone Medical Center's Genome Technology Center using a spike-in of 30\% PhiX DNA to control for low diversity in the library at the barcode and restriction sites.
+Sequences are available to download from the NCBI Short Read Archive (accession number \textcolor{red}{SRAXXXXXX.X}).
+
+%\begin{itemize}
+	%\item 6 animals, table with sources and other info
+	%\item Etter et al. 2011, though with modifications
+	%\item How we picked enzyme, PspXI. UCSC oligoMatch
+	%\item Adapter sequences. Barcodes with at least 3 mismatches. PE adapter
+	%\item What did it look like on BioA? Size? Concentration?
+	%\item Sequenced on Illumina MiSeq. 150PE
+	%\item NYU Langone Medical Center's Genome Technology Center (name right?)
+	%\item How much actually loaded?
+	%\item 30\% spike in with PhiX control DNA to control for low diversity library
+%\end{itemize}
 

### 2012-12-09 trunk/paper/paper.tex
Added sequence analysis to paper

@@
 
-%\begin{itemize}
-	%\item 6 animals, table with sources and other info
-	%\item Etter et al. 2011, though with modifications
-	%\item How we picked enzyme, PspXI. UCSC oligoMatch
-	%\item Adapter sequences. Barcodes with at least 3 mismatches. PE adapter
-	%\item What did it look like on BioA? Size? Concentration?
-	%\item Sequenced on Illumina MiSeq. 150PE
-	%\item NYU Langone Medical Center's Genome Technology Center (name right?)
-	%\item How much actually loaded?
-	%\item 30\% spike in with PhiX control DNA to control for low diversity library
-%\end{itemize}
+\paragraph{Sequence Analysis}
 
-\paragraph{Analysis Pipeline - Mapping to Reference Genomes}
-\begin{itemize}
-	\item Demultiplex. Must have barcode and restriction site intact.
-	\item Analyze reads with FastQC (No need to say this)
-	\begin{itemize}
-		\item Total sequence bp
-		\item Maximum possible sequence depth
-		\item Other stats that FastQC gives you
-	\end{itemize}
-	\item Aligned to rhesus genome using BWA aln
-	\begin{itemize}
-		\item default parameters
-	\end{itemize}
-	\item Combine paired-end reads with BWA sampe (No need to say this)
-	\item Convert to BAM, sort and index with samtools (No need to say this)
-	\item Analyze mapped reads with samtools utilities flagstat and idxstats and bamtools utility 
-	\item Post-alignment filtering steps
-	\begin{itemize}
-		\item Fix mate pair info with Picard
-		\item Filter for mapped and paired.
-		\item Remove dups with Picard
-		\item Add read group info with Picard (No need to say this)
-		\item Remove reads with low mapping quality with bamtools
-	\end{itemize}
-\end{itemize}
+	Sequence reads were demultiplexed, or separated by barcode, and reads without an expected barcode or an intact restriction enzyme cut site were excluded from the analysis.
+Reads were then aligned to the rhesus macaque reference genome (v.1.0, Mmul_051212/rheMac2, \textcolor{red}{ref}) using BWA with default parameters.
+Reads that were unmapped, unpaired, duplicates, or that had low mapping quality were removed after alignment using Picard (\textcolor{red}{ref}) and BamTools (\textcolor{red}{ref}).
 
-\paragraph{Analysis Pipeline - Variant Calling}
-\begin{itemize}
-	\item Local realignment with GATK
-	\item Fix paired end data with Picard (No need to say this)
-	\item Call SNPs with samtools. Min coverage = 3, max = 100
-	\item Summarize SNP stats with vcf-stats (No need to say this)
-	%\item Call consensus sequence with samtools
-\end{itemize}
+	After performing local realignment around indels with GATK (\textcolor{red}{ref}), SNPs and short indels were identified using SAMtools mpileup and BCFtools (\textcolor{red}{ref}).
+A minimum coverage of 3 reads and a maximum of 100 was required to call a SNP or an indel at a given location.
+
+	To analyze the degree of overlap between multiplexed individual's datasets, we...
 
@@
 \begin{itemize}
+	\item Info from analyzing reads with FastQC
+	\begin{itemize}
+		\item Number of reads (per ind. too see table)
+		\item Total sequence bp (per ind. too see table)
+		\item Maximum possible sequence depth (Cut?)
+		\item Other stats that FastQC gives you?
+	\end{itemize}
 	\item Table: 
@@
 		\item Number of reads per animal
+		\item Total sequenced bp per animal?
 		\item Number that passed filtration
@@
 	\end{itemize}
+	\item SNP info from merged analysis
 	\item SNP Venn diagram?

### 2012-12-09 trunk/paper/paper.tex
Added sample list table to paper

@@
 
+\begin{table}[h]
+\caption{Samples}
+\begin{center}
+  \begin{tabular}{ l | l | l }
+    \hline
+    Taxon & Source & Geographic Locality \\ \hline\hline
+    \textcolor{red}{Langur?} & NYU Collection & Unknown \\ \hline
+    \emph{Allenopithecus nigroviridis} & NYU Collection & Unknown \\ \hline
+    \emph{Macaca mulatta} & NYU Collection & Unknown \\ \hline
+    \emph{Cercocebus torquatus atys} & NYU Collection & Unknown \\ \hline
+    \emph{Papio anubis\textcolor{red}{?}} & NYU Collection & Unknown \\ \hline
+    \emph{Theropithecus gelada} & NYU Collection & Unknown \\ 
+    \hline
+  \end{tabular}
+\end{center}
+\end{table}
+
 \paragraph{Sequence Analysis}
@@
 	Sequence reads were demultiplexed, or separated by barcode, and reads without an expected barcode or an intact restriction enzyme cut site were excluded from the analysis.
-Reads were then aligned to the rhesus macaque reference genome (v.1.0, Mmul_051212/rheMac2, \textcolor{red}{ref}) using BWA with default parameters.
+Reads were then aligned to the rhesus macaque reference genome (v.1.0, Mmul\_051212/rheMac2, \textcolor{red}{ref}) using BWA with default parameters.
 Reads that were unmapped, unpaired, duplicates, or that had low mapping quality were removed after alignment using Picard (\textcolor{red}{ref}) and BamTools (\textcolor{red}{ref}).

### 2012-12-09 paper/paper.tex
Started on results section in paper

--- 2012-12-09 trunk/paper/paper.tex
@@
 
-	Genomic DNA from 6 animals was digested with \emph{PspXI} (New England Biolabs) and used to create a multiplexed RAD tag library.
+	Genomic DNA from 6 primates was digested with \emph{PspXI} (New England Biolabs) and used to create a multiplexed RAD tag library.
 Our library preparation method followed that of Etter et al, 2011 with the following modifications: the P1 adapter top\textcolor{red}{(?)} oligonucleotide was modified to have an overhang corresponding to the cut site of \emph{PspXI}, and a longer P2 adapter suitable for paired end sequencing was used (P2\_top: 5'-\textcolor{red}{SEQUENCEHERE}-3'; P2\_bottom: 5'-\textcolor{red}{SEQUENCEHERE}-3').
@@
 We sequenced the prepared library as one 150-cycle paired-end run of an Illumina MiSeq at the NYU Langone Medical Center's Genome Technology Center using a spike-in of 30\% PhiX DNA to control for low diversity in the library at the barcode and restriction sites.
+Other individuals were sequenced alongside those of the present study.
 Sequences are available to download from the NCBI Short Read Archive (accession number \textcolor{red}{SRAXXXXXX.X}).
@@
 A minimum coverage of 3 reads and a maximum of 100 was required to call a SNP or an indel at a given location.
+Orthologous SNPs were tallied using VCFtools (\textcolor{red}{ref}).
 
-	To analyze the degree of overlap between multiplexed individual's datasets, we...
-
-\paragraph{Analysis Pipeline - Analysis of Degree of Overlap}
-\begin{itemize}
-	\item Calculate coverage of restriction site-associated regions
-	\begin{itemize}
-		\item Info on targeted intervals
-		\begin{itemize}
-			\item Total number possible targets in rhesus genome (compare to human too?)
-			\item Total possible target BP
-		\end{itemize}
-		\item How many targets did we hit?
-		\begin{itemize}
-			\item BEDtools multiBamCoverage for this job
-			\item Number and percentage of targets with coverage $\ge 1$
-			\item Number and percentage of targets with coverage $\ge N$
-		\end{itemize}
-	\end{itemize}
-	\item Count orthologous SNPs shared between individuals
-	\begin{itemize}
-		\item VCFtools vcf-compare for this job
-	\end{itemize}
-\end{itemize}
+	To assess how many restriction sites were successfully sequenced and to analyze the degree of overlap between multiplexed individual's datasets, we first found all possible \emph{PspXI} cut sites using the oligoMatch utility in the USCS Genome Browser program (\textcolor{red}{ref}).
+This allowed us to calculate the coverage of these restriction site-associated regions using BEDtools' multiBamCov program (\textcolor{red}{ref}).
 
@@
 \section{Results}
+
+% 3804227030 bp total (12,983,710 pairs of reads * (142bp + 151bp). This includes non-OWM taxa, remember.
+% OWM sequencing reads: 6,125,860 total, half that many pairs
+% Of those, 2029586 passed the filtration steps
+
+6.1 million sequencing reads with an intact barcode and restriction enzyme cut site could be assigned confidently to one of the six Old World monkeys in the present study.
+Roughly 2.0 million of those reads were successfully mapped to the rhesus macaque genome and passed all quality control filtration steps.
+Relative to the reference genome, our study identified 531,175 SNPs and 24,260 small indels among all samples.
+Information for each individual is summarized in Table 1.
+
+In the rhesus macaque genome, we found 108,727 possible cut sites for \emph{PspXI}.
+
+\begin{table}[h]
+\caption{Individual Sample Information}
+\begin{center}
+	\small
+	\begin{tabular}{ p{3cm} || l || p{1.75cm} | p{1.75cm} || p{1.75cm} | p{1.75cm} | l }
+		\hline
+		Taxon & Source & \# Reads & \# Filtered Reads & \# Loci $\ge 1$ Read & \# Loci $\ge 3$ Reads & \# SNPs \\ \hline\hline
+		\textcolor{red}{Langur?} & \textcolor{red}{Unknown} & 1,367,050 & 306,248 & & & 363,501 \\ \hline
+		\emph{Allenopithecus nigroviridis} & \textcolor{red}{Unknown} & 1,325,558 & 447,279 & & & 356,750 \\ \hline
+		\emph{Macaca mulatta} & \textcolor{red}{Unknown} & 481,376 & 234,659 & & & 279,350 \\ \hline
+		\emph{Cercocebus torquatus atys} & \textcolor{red}{Unknown} & 297,290 & 124,700 & & & 319,766 \\ \hline
+		\emph{Papio anubis\textcolor{red}{?}} & \textcolor{red}{Unknown} & 743,556 & 313,676 & & & 316,118 \\ \hline
+		\emph{Theropithecus gelada} & \textcolor{red}{Unknown} & 1,911,030 & 603,024 & & & 366,097 \\
+		\hline
+	\end{tabular}
+\end{center}
+\end{table}
+
 \begin{itemize}
-	\item Info from analyzing reads with FastQC
-	\begin{itemize}
-		\item Number of reads (per ind. too see table)
-		\item Total sequence bp (per ind. too see table)
-		\item Maximum possible sequence depth (Cut?)
-		\item Other stats that FastQC gives you?
-	\end{itemize}
-	\item Table: 
-	\begin{itemize}
-		\item Number of reads per animal
-		\item Total sequenced bp per animal?
-		\item Number that passed filtration
 	\item Number of loci hit
 	\item Number of loci hit with coverage $\ge N$
-		\item Number of SNPs
-	\end{itemize}
 	\item SNP info from merged analysis
 	\item SNP Venn diagram?
-	\item Table of overlapping region, orthologous SNP counts
-	%\item Phylogenetic tree
+	\item Calculate coverage of restriction site-associated regions
+	\item Total number possible targets in rhesus genome (compare to human too?)
+	\item Total possible target BP
+	\item How many targets did we hit?
+	\item BEDtools multiBamCoverage for this job
+	\item Number and percentage of targets with coverage $\ge 1$
+	\item Number and percentage of targets with coverage $\ge N$
+	\item Count orthologous SNPs shared between individuals
+	\item VCFtools vcf-compare for this job
 \end{itemize}

### 2012-12-09 paper/paper.tex
Results done up through coverage analysis. No SNPs yet

@@
 			\and
-			Todd, I suppose
+			Todd R. Disotell
 	}
@@
 \usepackage{color}
+\usepackage{graphics}
 
@@
 
-In the rhesus macaque genome, we found 108,727 possible cut sites for \emph{PspXI}.
-
 \begin{table}[h]
@@
 		Taxon & Source & \# Reads & \# Filtered Reads & \# Loci $\ge 1$ Read & \# Loci $\ge 3$ Reads & \# SNPs \\ \hline\hline
-		\textcolor{red}{Langur?} & \textcolor{red}{Unknown} & 1,367,050 & 306,248 & & & 363,501 \\ \hline
-		\emph{Allenopithecus nigroviridis} & \textcolor{red}{Unknown} & 1,325,558 & 447,279 & & & 356,750 \\ \hline
-		\emph{Macaca mulatta} & \textcolor{red}{Unknown} & 481,376 & 234,659 & & & 279,350 \\ \hline
-		\emph{Cercocebus torquatus atys} & \textcolor{red}{Unknown} & 297,290 & 124,700 & & & 319,766 \\ \hline
-		\emph{Papio anubis\textcolor{red}{?}} & \textcolor{red}{Unknown} & 743,556 & 313,676 & & & 316,118 \\ \hline
-		\emph{Theropithecus gelada} & \textcolor{red}{Unknown} & 1,911,030 & 603,024 & & & 366,097 \\
+		\textcolor{red}{Langur?} & \textcolor{red}{Unknown} & 1,367,050 & 306,248 & 47,516 & 24,052 & 363,501 \\ \hline
+		\emph{Allenopithecus nigroviridis} & \textcolor{red}{Unknown} & 1,325,558 & 447,279 & 62,824 & 40,244 & 356,750 \\ \hline
+		\emph{Macaca mulatta} & \textcolor{red}{Unknown} & 481,376 & 234,659 & 77,798 & 32,302 & 279,350 \\ \hline
+		\emph{Cercocebus torquatus atys} & \textcolor{red}{Unknown} & 297,290 & 124,700 & 48,440 & 13,104 & 319,766 \\ \hline
+		\emph{Papio anubis\textcolor{red}{?}} & \textcolor{red}{Unknown} & 743,556 & 313,676 & 74,582 & 39,062 & 316,118 \\ \hline
+		\emph{Theropithecus gelada} & \textcolor{red}{Unknown} & 1,911,030 & 603,024 & 68,440 & 50,656 & 366,097 \\
 		\hline
@@
 
+In the rhesus macaque genome, we found 54,364 possible cut sites for \emph{PspXI} and 108,728 possible sequencing sites (two per cut site, one upstream and one downstream).
+Of those, 101,138 sites (93.02\%) were covered by at least one read in at least one individual, and for 13,456 sites (12.38\%), all six individuals had at least one read (Figure 1).
+When we restrict the analysis to sites with at least three reads, 80,448 sites (73.99\%) were covered in at least one individual and 1,354 sites (1.25\%) had all six individuals present.
+
+\includegraphics{figs/seq_site_coverage_by_ind}
+
 \begin{itemize}
-	\item Number of loci hit
-	\item Number of loci hit with coverage $\ge N$
 	\item SNP info from merged analysis
 	\item SNP Venn diagram?
-	\item Calculate coverage of restriction site-associated regions
-	\item Total number possible targets in rhesus genome (compare to human too?)
-	\item Total possible target BP
-	\item How many targets did we hit?
-	\item BEDtools multiBamCoverage for this job
-	\item Number and percentage of targets with coverage $\ge 1$
-	\item Number and percentage of targets with coverage $\ge N$
 	\item Count orthologous SNPs shared between individuals

### 2012-12-09 trunk/paper/paper.tex
Tweaked things a bit before leaving for the night

--- 2012-12-09 trunk/paper/paper.tex
@@
 \section{Results}
+
+% 3804227030 bp total (12,983,710 pairs of reads * (142bp + 151bp). This includes non-OWM taxa, remember.
+% OWM sequencing reads: 6,125,860 total, half that many pairs
+% Of those, 2029586 passed the filtration steps
+
+6.1 million sequencing reads with an intact barcode and restriction enzyme cut site could be assigned confidently to one of the six Old World monkeys in the present study.
+Roughly 2.0 million of those reads were successfully mapped to the rhesus macaque genome and passed all quality control filtration steps.
+Relative to the reference genome, our study identified 531,175 SNPs and 24,260 small indels among all samples.
+Information for each individual is summarized in Table 1.
+
+\begin{table}[h]
+\caption{Individual Sample Information}
+\begin{center}
+	\small
+	\begin{tabular}{ p{3cm} || l || p{1.75cm} | p{1.75cm} || p{1.75cm} | p{1.75cm} | l }
+		\hline
+		Taxon & Source & \# Reads & \# Filtered Reads & \# Loci $\ge 1$ Read & \# Loci $\ge 3$ Reads & \# SNPs \\ \hline\hline
+		\textcolor{red}{Langur?} & \textcolor{red}{Unknown} & 1,367,050 & 306,248 & 47,516 & 24,052 & 363,501 \\ \hline
+		\emph{Allenopithecus nigroviridis} & \textcolor{red}{Unknown} & 1,325,558 & 447,279 & 62,824 & 40,244 & 356,750 \\ \hline
+		\emph{Macaca mulatta} & \textcolor{red}{Unknown} & 481,376 & 234,659 & 77,798 & 32,302 & 279,350 \\ \hline
+		\emph{Cercocebus torquatus atys} & \textcolor{red}{Unknown} & 297,290 & 124,700 & 48,440 & 13,104 & 319,766 \\ \hline
+		\emph{Papio anubis\textcolor{red}{?}} & \textcolor{red}{Unknown} & 743,556 & 313,676 & 74,582 & 39,062 & 316,118 \\ \hline
+		\emph{Theropithecus gelada} & \textcolor{red}{Unknown} & 1,911,030 & 603,024 & 68,440 & 50,656 & 366,097 \\
+		\hline
+	\end{tabular}
+\end{center}
+\end{table}
+
+In the rhesus macaque genome, we found 54,364 possible cut sites for \emph{PspXI} and 108,728 possible sequencing sites (two per cut site, one upstream and one downstream).
+Of those, 101,138 sites (93.02\%) were covered by at least one read in at least one individual, and for 13,456 sites (12.38\%), all six individuals had at least one read (Figure 1).
+When we restrict the analysis to sites with at least three reads, 80,448 sites (73.99\%) were covered in at least one individual and 1,354 sites (1.25\%) had all six individuals present.
+
 \begin{itemize}
-	\item Info from analyzing reads with FastQC
+	\item Pairwise comparisons between individuals? Like X orthologous regions shared between Papio and Thero?
+\end{itemize}
+
+\includegraphics{figs/seq_site_coverage_by_ind}
+
 \begin{itemize}
-		\item Number of reads (per ind. too see table)
-		\item Total sequence bp (per ind. too see table)
-		\item Maximum possible sequence depth (Cut?)
-		\item Other stats that FastQC gives you?
-	\end{itemize}
-	\item Table: 
-	\begin{itemize}
-		\item Number of reads per animal
-		\item Total sequenced bp per animal?
-		\item Number that passed filtration
-		\item Number of loci hit
-		\item Number of loci hit with coverage $\ge N$
-		\item Number of SNPs
-	\end{itemize}
 	\item SNP info from merged analysis
 	\item SNP Venn diagram?
-	\item Table of overlapping region, orthologous SNP counts
-	%\item Phylogenetic tree
+	\item Count orthologous SNPs shared between individuals. Pairwise?
 \end{itemize}
@@
 	\item RADseq is viable tool for researcher interested in primate phylogenetics, pop. gen.
+	\begin{itemize}
+		\item Cheap and easy to create libraries. Competent lab can do it in two days. (Ha!)
+		\item Multiplexing is great for pop gen studies.
 		\item Enzyme choice allows control over coverage, number of individuals, number of loci.
+	\end{itemize}
 	\item Potential problems with RADseq method
@@
 \section{Acknowledgements}
-Acknowledgements
+\begin{itemize}
+	\item NYU Med School folk
+	\item \textcolor{red}{Grants?}
+\end{itemize}
 

### 2012-12-10 paper/paper.tex
Changed pipeline to incorporate SE data.

@@
 			\and
-			Andrew S. Burrell
-			\and
-			Luca S.J. Pozzi
+			Luca Pozzi
 			\and
 			Todd R. Disotell
+			\and
+			Andrew S. Burrell
 	}
@@
 \begin{abstract}
-\ldots Blah, blah, blah, RADseq, blah, blah, Cercopithecoidea. \ldots
+\ldots (This is the blurb from the email.) Our paper is an introduction to a 2nd generation sequencing technique for typing thousands of genome-wide markers from non-model organisms.
+Though it's been used in other taxa, this would be the first published application to primates.
+We demonstrate it with six Cercopithecoids and discuss its promise for doing mutli-locus population genetics in primates.
+\ldots
 \end{abstract}
@@
 
-%\paragraph{Analysis Pipeline - Inferring Phylogeny}
-%\begin{itemize}
-%	\item Using Stacks? 
-%	\item Using method like cichlid people?
-%	\item Using method like Rubin et al %http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0033394
-%\end{itemize}
+\paragraph{Analysis Pipeline - Inferring Phylogeny}
+\begin{itemize}
+	\item Using method like cichlid people?
+	\item Using method like Rubin et al? 
+	\item Concatenated SNPs (not indels) into alignment. 
+	\item Dimensions = 6 x ???
+\end{itemize}
 
@@
 		Taxon & Source & \# Reads & \# Filtered Reads & \# Loci $\ge 1$ Read & \# Loci $\ge 3$ Reads & \# SNPs \\ \hline\hline
-		\textcolor{red}{Langur?} & \textcolor{red}{Unknown} & 1,367,050 & 306,248 & 47,516 & 24,052 & 363,501 \\ \hline
+		\emph{Semnopithecus entellus} & \textcolor{red}{Unknown} & 1,367,050 & 306,248 & 47,516 & 24,052 & 363,501 \\ \hline
 		\emph{Allenopithecus nigroviridis} & \textcolor{red}{Unknown} & 1,325,558 & 447,279 & 62,824 & 40,244 & 356,750 \\ \hline
@@
 
+\begin{itemize}
+	\item Pairwise comparisons between individuals? Like X orthologous regions shared between Papio and Thero?
+\end{itemize}
+
 \includegraphics{figs/seq_site_coverage_by_ind}
@@
 	\item SNP Venn diagram?
-	\item Count orthologous SNPs shared between individuals
-	\item VCFtools vcf-compare for this job
+	\item Count orthologous SNPs shared between individuals. Pairwise?
 \end{itemize}

### 2012-12-10 scripts/alignSE.sh

@@ -1,7 +1,7 @@
 #!/bin/sh
 
 # ------------------------------------------------------------------------------
-# --- Align reads to genome with BWA
+# --- Align SE reads to genome with BWA. Then sneak them into PE BAM
 # ------------------------------------------------------------------------------
 
 # Check that genome FASTA, and genome code were passed as parameters
@@ -18,21 +18,13 @@ GENOME_CODE=$2
 echo "CMD: $BWA/bwa aln \
 	$BWA_ALN_PARAM \
 	$GENOME_PATH \
-	$READ1 > results/${IND_ID}.read1.bwa.${GENOME_CODE}.sai";
+	$READ_SE \
+	> results/${IND_ID}.readSE.bwa.${GENOME_CODE}.sai;";
 
 $BWA/bwa aln \
 	$BWA_ALN_PARAM \
 	$GENOME_PATH \
-	$READ1 > results/${IND_ID}.read1.bwa.${GENOME_CODE}.sai
-
-echo "CMD: $BWA/bwa aln \
-	$BWA_ALN_PARAM \
-	$GENOME_PATH \
-	$READ2 > results/${IND_ID}.read2.bwa.${GENOME_CODE}.sai";
-
-$BWA/bwa aln \
-	$BWA_ALN_PARAM \
-	$GENOME_PATH \
-	$READ2 > results/${IND_ID}.read2.bwa.${GENOME_CODE}.sai
+	$READ_SE \
+	> results/${IND_ID}.readSE.bwa.${GENOME_CODE}.sai;
 
 exit;

### 2012-12-10 scripts/filter_mapped_reads_paired.sh

@@ -13,11 +13,10 @@ fi
 
 GENOME_CODE=$1
 
-# Removed "-isProperPair true"
+# Removed "-isProperPair true" and "-isPaired true"
 
 ${BAMTOOLS}/bamtools filter \
 	-isMapped true \
-	-isPaired true \
 	-in results/${IND_ID}.bwa.${GENOME_CODE}.fixed.bam \
 	-out results/${IND_ID}.bwa.${GENOME_CODE}.fixed.filtered.bam
 

### 2012-12-10 scripts/filter_mapped_reads_quality.sh

@@ -15,7 +15,7 @@ GENOME_CODE=$1
 
 ${BAMTOOLS}/bamtools filter \
 	-mapQuality ">=20" \
-	-in results/${IND_ID}.bwa.${GENOME_CODE}.fixed.filtered.nodup.RG.bam \
+	-in results/${IND_ID}.bwa.${GENOME_CODE}.fixed.filtered.RG.bam \
 	-out results/${IND_ID}.bwa.${GENOME_CODE}.passed.bam
 
 exit;

### 2012-12-10 scripts/fix_mate_pairs.sh

@@ -21,7 +21,7 @@ mkdir -p $TMP_DIR
 # Also shorten the filename
 java -Djava.io.tmpdir=${TMP_DIR} \
 	-jar ${PICARD}/FixMateInformation.jar \
-	INPUT= results/${IND_ID}.bwa.${GENOME_CODE}.sam.bam.sorted.bam \
+	INPUT= results/${IND_ID}_MERGED.bwa.${GENOME_CODE}.sam.bam.sorted.bam \
 	OUTPUT=results/${IND_ID}.bwa.${GENOME_CODE}.fixed.bam \
 	SO=coordinate \
 	VALIDATION_STRINGENCY=LENIENT \

### 2012-12-10 trunk/scripts/add_read_groups.sh
Changed pipeline to incorporate SE data.

@@ -16,8 +16,8 @@ GENOME_CODE=$1
 # Add read groups with Picard:
 
 java -jar ${PICARD}/AddOrReplaceReadGroups.jar \
-	INPUT=results/${IND_ID}.bwa.${GENOME_CODE}.fixed.filtered.nodup.bam \
-	OUTPUT=results/${IND_ID}.bwa.${GENOME_CODE}.fixed.filtered.nodup.RG.bam \
+	INPUT=results/${IND_ID}.bwa.${GENOME_CODE}.fixed.filtered.bam \
+	OUTPUT=results/${IND_ID}.bwa.${GENOME_CODE}.fixed.filtered.RG.bam \
 	RGLB=${IND_ID} \
 	RGPL=Illumina \
 	RGPU=Group1 \

### 2012-12-10 trunk/scripts/analyze_enzyme_count_data.R

@@ -1,0 +1,80 @@
+
+rad_counts=read.table("RAD_coverage.txt")
+names(rad_counts)=c("chr","start","end", "enz", "len", "str", "Allenopithecus", "Coast_langur", "CTA", "Mmulatta1", "Papio19349", "Tgelada871162")
+
+rad_counts_only = rad_counts[,7:12]
+
+rad_present = rad_counts_only > 0
+rad_gt3     = rad_counts_only >= 3
+
+rad_sums = rowSums(rad_present)
+rad_sums_gt3 = rowSums(rad_gt3)
+
+#	length(rad_sums[rad_sums == 6]) # 13456
+#	length(rad_sums[rad_sums == 5]) # 23350
+#	length(rad_sums[rad_sums == 4]) # 22890
+#	length(rad_sums[rad_sums == 3]) # 18488
+#	length(rad_sums[rad_sums == 2]) # 12136
+#	length(rad_sums[rad_sums == 1]) # 10818
+#	length(rad_sums[rad_sums == 0]) # 7590
+#	
+#	length(rad_sums[rad_sums >= 6]) # 13456
+#	length(rad_sums[rad_sums >= 5]) # 36806
+#	length(rad_sums[rad_sums >= 4]) # 59696
+#	length(rad_sums[rad_sums >= 3]) # 78184
+#	length(rad_sums[rad_sums >= 2]) # 90320
+#	length(rad_sums[rad_sums >= 1]) # 101138
+#	length(rad_sums[rad_sums >= 0]) # 108728
+
+x = 0:6
+
+### One or more reads --------------------------------------------------------------------
+
+counts = lapply(x, get_counts <- function(c) length(rad_sums[rad_sums == c]))
+counts = c(do.call("cbind",counts)) 
+counts
+pdf("~/RADseq_primates/RADprimates/paper/figs/seq_site_coverage_by_ind.pdf", height=4, width=4)
+barplot(counts, 
+	ylab="Count of Possible Sequencing Sites", 
+	xlab="Number of Individuals with at Least One Read",
+	names=x,
+	col=1)
+dev.off()
+
+cumu_counts = lapply(x, get_cumu_counts <- function(c) length(rad_sums[rad_sums >= c]))
+cumu_counts = c(do.call("cbind",cumu_counts)) 
+cumu_counts
+plot(cumu_counts ~ x)
+
[…]

### 2012-12-10 trunk/scripts/call_snps.sh

@@ -12,17 +12,19 @@ if [ -z "$2" ]; then
 fi
 
 IN_BAM=$1
-OUT_BCF=$(echo $IN_BAM | sed 's/\.bam/.raw.bcf/')
+OUT_BCF=$(echo $IN_BAM | sed 's/\.bam$/.raw.bcf/')
 GENOME_FA=$2
 
 echo "${SAMTOOLS}/samtools mpileup \
-	-uf ${GENOME_FA} \
+	-C50 \
+	-Auf ${GENOME_FA} \
 	${IN_BAM}  | \
 	${BCFTOOLS}/bcftools view -bvcg - \
 	> ${OUT_BCF}";
 
 ${SAMTOOLS}/samtools mpileup \
-	-uf ${GENOME_FA} \
+	-C50 \
+	-Auf ${GENOME_FA} \
 	${IN_BAM}  | \
 	${BCFTOOLS}/bcftools view -bvcg - \
 	> ${OUT_BCF}

### 2012-12-10 trunk/scripts/convert_fasta_to_nexus.pl

@@ -1,0 +1,20 @@
+#!/usr/bin/perl
+
+# Program to convert FASTA alignment output from vcf_tab_to_fasta_alignment.pl to
+# NEXUS format.
+
+use strict;
+use warnings;
+
+use lib '/home/cmb433/local_perl/';
+
+use Bio::AlignIO;
+
+my $in  = Bio::AlignIO->new(	-file   => "results/merged.align.fasta",
+								-format => "fasta");
+my $out = Bio::AlignIO->new(	-file => ">results/merged.align.nex",
+								-format => "nexus");
+ 
+while (my $aln = $in->next_aln) { 
+	$out->write_aln($aln); 
+}

### 2012-12-10 trunk/scripts/convert_fasta_to_phylip.pl

@@ -1,0 +1,20 @@
+#!/usr/bin/perl
+
+# Program to convert FASTA alignment output from vcf_tab_to_fasta_alignment.pl to
+# PHYLIP format.
+
+use strict;
+use warnings;
+
+use lib '/home/cmb433/local_perl/';
+
+use Bio::AlignIO;
+
+my $in  = Bio::AlignIO->new(	-file   => "results/merged.align.fasta",
+								-format => "fasta");
+my $out = Bio::AlignIO->new(	-file => ">results/merged.align.phy",
+								-format => "phylip");
+ 
+while (my $aln = $in->next_aln) { 
+	$out->write_aln($aln); 
+}

### 2012-12-10 trunk/scripts/sam2bam.sh

@@ -4,22 +4,26 @@
 # --- Convert SAM to BAM
 # ------------------------------------------------------------------------------
 
-# Check that genome FASTA and genome code were passed as parameters
-USAGE="$0 genome.index.fai genome_code";
+# Check that genome FASTA and input SAM were passed as parameters
+USAGE="$0 genome.index.fai in.sam";
 if [ -z "$2" ]; then
 	echo "ERROR: $USAGE";
 	exit 1;
 fi
 
 INDEX_FAI=$1
-GENOME_CODE=$2
+IN_SAM=$2
+
+echo "$SAMTOOLS/samtools view \
+	-b \
+	-t ${INDEX_FAI} \
+	-o ${IN_SAM}.bam \
+	$IN_SAM";
 
 $SAMTOOLS/samtools view \
 	-b \
 	-t ${INDEX_FAI} \
-	-o results/${IND_ID}.bwa.${GENOME_CODE}.sam.bam \
-	results/${IND_ID}.bwa.${GENOME_CODE}.sam
-
-echo results/${IND_ID}.bwa.${GENOME_CODE}.sam
+	-o ${IN_SAM}.bam \
+	$IN_SAM
 
 exit;

### 2012-12-10 trunk/scripts/vcf_tab_to_fasta_alignment.pl

@@ -1,0 +1,155 @@
+#!/usr/bin/perl
+
+# Program to convert output of VCFtools' vcf-to-tab (modified to only export IUPAC SNPs)
+# to FASTA alignment.
+
+# Sample input file
+#	$ head results/merged.flt.vcf.tab
+#	chr10	94051	C	./	./	./	./	./	T/T
+#	chr10	94056	T	./	./	./	./	./	C/C
+#	chr10	94180	G	./	A/A	./	./	./	./
+
+
+use strict;
+use warnings;
+
+my %iupac = (
+			'G/G' => 'G',
+			'C/C' => 'C',
+			'T/T' => 'T',
+			'A/A' => 'A',
+
+			'G/T' => 'K',
+			'T/G' => 'K',
+			'A/C' => 'M',
+			'C/A' => 'M',
+			'C/G' => 'S',
+			'G/C' => 'S',
+			'A/G' => 'R',
+			'G/A' => 'R',
+			'A/T' => 'W',
+			'T/A' => 'W',
+			'C/T' => 'Y',
+			'T/C' => 'Y',
+
+			'./.' => '.',
+		);
+
+my $input_tab = "results/merged.flt.vcf.tab";
+
+open (TAB, "<$input_tab")
+	or die "ERROR: Could not open input file $input_tab.\n";
+
+my $header = <TAB>;
+
+my @col_names = split /\t/, $header;
+
+# Make temporary file with just lines we're going to use
+my $temp_tab = $input_tab . "_clean";
+open (TEMP, ">$temp_tab")
+	or die "ERROR: Could not open temp file $temp_tab.\n";
+
+LINE: foreach my $line (<TAB>) {
+
+	my @data = split /\t/, $line;
+	
+	# Skip if this is indel (Length of @data will be less than 9)
+	if ((scalar @data) < 9) {
+		next LINE;
+	}
+	
+	# Skip if any basepairs are actually 2 or more together
+	for (my $i = 2; $i < 9; $i++) {
+		
+		my $bp = $data[$i]; 
+		chomp $bp;
+		if ($bp =~ /\w{2,}/) {
+			next LINE;
+		}
+	}
+	
+	# Exclude heterozygotes. Keep only fixed SNPs
+	for (my $i = 2; $i < 9; $i++) {
+		
+		my $bp = $data[$i]; 
+		chomp $bp;
+		if ($bp =~ /(\w)\/(\w)/) {
+			if ($1 ne $2) {
+				next LINE;
+			}
+		}
+	}
+	
[…]

### 2012-12-10 trunk/scripts/add_read_groups.sh
Starting to add comparison steps to Makefile

@@ -21,6 +21,6 @@ java -jar ${PICARD}/AddOrReplaceReadGroups.jar \
 	RGLB=${IND_ID} \
 	RGPL=Illumina \
 	RGPU=Group1 \
-	RGSM=Grp1
+	RGSM=${IND_ID}
 
 exit;

### 2012-12-10 trunk/scripts/filter_snps.sh

@@ -22,5 +22,12 @@ ${BCFTOOLS}/bcftools view ${IN_BCF} | \
 	${BCFTOOLS}/vcfutils.pl varFilter -d ${SNP_MIN_COV} -D ${SNP_MAX_COV} \
 	> ${OUT_VCF};
 
+# While we're here, make indexed *.vcf.gz files
+
+export PATH=$PATH:$TABIX
+
+bgzip -c ${OUT_VCF} > ${OUT_VCF}.gz
+tabix -p vcf ${OUT_VCF}.gz
+
 exit;
 

### 2012-12-10 trunk/scripts/merge_vcf.sh

@@ -1,0 +1,15 @@
+#!/bin/sh
+
+# ------------------------------------------------------------------------------
+# --- Merge VCF files from individual SNP calling
+# ------------------------------------------------------------------------------
+
+export PATH=$PATH:$TABIX
+
+ALL_VCF_GZ=$(ls results/*.bwa.rhesus.passed.realn.bam.RG.flt.vcf.gz)
+
+${VCFTOOLS}/vcf-merge \
+	${ALL_VCF_GZ} \
+	> results/merged.flt.vcf
+
+exit;

### 2012-12-12 trunk/paper/paper.tex
Started intro. Switching over from Cercopithecoids to across Primates.

@@
 \section{Introduction}
-\begin{itemize}
-	\item Next-gen sequencing revolution promises gains in primatology
-	\item Still expensive
-	\item Many genomes, but still tough doing genomics on non-model organisms
-	\item What is RADseq?
-	\item Previous RADseq studies
-	\item Why would it be good for primates
-	\item PRESENT STUDY
-	\begin{itemize}
-		\item We did RADseq on 6 Cercopithecoids
-		\item Assessed how well it worked
-		\item Show it has promise for primates
-	\end{itemize}
-\end{itemize}
+
+During the first decades of molecular primatology, a formidable impediment to researchers was the need to develop and type polymorphic markers in a taxon of interest.
+The markers that resulted from this tedious and expensive task were often uninformative when applied outside of the population used in their design, necessitating further rounds of primer design or microsatellite assays, for example (Davey et al., 2011).
+Due to the bottleneck of marker discovery, many population genetic or phylogenetic studies in molecular primatology have been based on one or few loci (Ting and Sterner, published yet?).
+Such inferences can reliably give the evolutionary history of those particular regions of the genome, but they fail to adequately capture the complete complex history of the population given the mosaic nature of genomic evolution (some gene-tree species tree review ref).
+Adequate resolution depends on high marker density, and until recently that goal has been out of reach for many primate researchers.
+
+The rapidly decreasing costs of DNA sequencing technology have promised revolutionary gains for primatology (Enard and Paabo 2004, maybe?
+Goodman et al 2005?).
+A primate researcher benefits from the many nearby sequenced and assembled reference genomes in the order, but genomic studies of non-model organisms nevertheless remain difficult.
+Though the cost of low coverage whole genome sequencing has fallen to a level feasible for many researchers' budgets, sequencing whole genomes for the tens or hundreds of individuals desired in a typical population genetic study is often prohibitively expensive and quite possibly superfluous.
+
+Fortunately, researchers have recently developed techniques that reduce the complexity of the genome and allow discovery and typing of thousands or tens of thousands of genome-wide makers in many individuals in a single step.
+RAD-seq is one such simple, inexpensive reduced representation technique which allows for the sequencing of small fragments of the genome adjacent to restriction enzyme cut sites (Baird et al., 2008).
+These RAD tags, or restriction-site association DNA tags, were originally developed for use in microarray hybridization typing (Miller et al., 2007), but an updated protocol substitutes second-generation DNA sequencing to rapidly discover and type SNPs (Baird et al., 2008).
+The lack of reliance on a reference genome and applicability to datasets of many individuals make it a promising technique for phylogenetic or population genetic studies in non-model organisms, such as many primates.
+
+In the present study, we summarize the RAD-seq method in brief and note the many and varied applications since its development.
+We go on to discuss reasons that make RAD-seq promising for primatologists, and then demonstrate the technique in 5 primates: a lemur, New World monkey, Old World monkey, and two apes.
+
+\subsection{The RAD-seq Technique}
+
+[The RAD-seq method, in brief. Also, Fig. 1]
+
+\subsection{Previous RAD-seq Studies}
+
+[Previous RADseq studies. Focus on pop gen, phylogeny studies]
+
+\subsection{Prospective Primate Applications}
+
+[Why it would be good in primates.]
+
+Changing the restriction enzyme to a more or less frequent cutter allows fine grained control over the number of loci, number of individuals multiplexed, and depth of coverage, allowing researchers to tailor marker discovery to their hypothesis.
 
@@
 
-	Genomic DNA from 6 primates was digested with \emph{PspXI} (New England Biolabs) and used to create a multiplexed RAD tag library.
+	Genomic DNA from 5 primates was digested with \emph{PspXI} (New England Biolabs) and used to create a multiplexed RAD tag library.
 Our library preparation method followed that of Etter et al, 2011 with the following modifications: the P1 adapter top\textcolor{red}{(?)} oligonucleotide was modified to have an overhang corresponding to the cut site of \emph{PspXI}, and a longer P2 adapter suitable for paired end sequencing was used (P2\_top: 5'-\textcolor{red}{SEQUENCEHERE}-3'; P2\_bottom: 5'-\textcolor{red}{SEQUENCEHERE}-3').
@@
 		\item Enzyme choice allows control over coverage, number of individuals, number of loci.
+		\item Baird: "different marker densities can be attained by choice of restriction enzyme"
 	\end{itemize}
 	\item Potential problems with RADseq method
+	\item ddRADseq
 	\item Promise for primatology

### 2012-12-12 paper/paper.tex
Wrote more of intro. Switched methods, results over to primate sample

--- 2012-12-09 paper/paper.tex
@@
 			\and
+			Luca Pozzi
+			\and
+			Todd R. Disotell
+			\and
 			Andrew S. Burrell
-			\and
-			Luca S.J. Pozzi
-			\and
-			Todd, I suppose
 	}
@@
 \begin{abstract}
-\ldots Blah, blah, blah, RADseq, blah, blah, Cercopithecoidea. \ldots
+\textcolor{red}{\ldots (This is just a bit of the blurb from the email.) Our paper is an introduction to a 2nd generation sequencing technique for typing thousands of genome-wide markers from non-model organisms.
+We demonstrate it in five primates and discuss its promise for doing mutli-locus phylogenetics and population genetics in primates.
+\ldots}
 \end{abstract}
@@
 \section{Introduction}
-\begin{itemize}
-	\item Next-gen sequencing revolution promises gains in primatology
-	\item Still expensive
-	\item Many genomes, but still tough doing genomics on non-model organisms
-	\item What is RADseq?
-	\item Previous RADseq studies
-	\item Why would it be good for primates
-	\item PRESENT STUDY
-	\begin{itemize}
-		\item We did RADseq on 6 Cercopithecoids
-		\item Assessed how well it worked
-		\item Show it has promise for primates
-	\end{itemize}
-\end{itemize}
+
+During the first decades of molecular primatology, a formidable impediment to researchers was the need to develop and type polymorphic markers in a taxon of interest.
+The markers that resulted from this tedious and expensive task were often uninformative when applied outside of the population used in their design, necessitating further rounds of primer design or microsatellite assays, for example (Davey et al., 2011).
+Due to the bottleneck of marker discovery, many population genetic or phylogenetic studies in molecular primatology have been based on one or few loci (\textcolor{red}{Ting and Sterner, published yet?}).
+Such inferences can reliably give the evolutionary history of those particular regions of the genome, but they fail to adequately capture the complete complex history of the population given the mosaic nature of genomic evolution (\textcolor{red}{some gene-tree species tree review ref}).
+Adequate resolution depends on high marker density, and until recently that goal has been out of reach for many primate researchers.
+
+The rapidly decreasing costs of DNA sequencing technology have promised revolutionary gains for primatology (\textcolor{red}{Enard and Paabo 2004, maybe?
+Goodman et al 2005?}).
+A primate researcher benefits from the many nearby sequenced and assembled reference genomes in the order, but genomic studies of non-model organisms nevertheless remain difficult.
+Though the cost of low coverage whole genome sequencing has fallen to a level feasible for many researchers' budgets, sequencing whole genomes for the tens or hundreds of individuals desired in a typical population genetic study is often prohibitively expensive and quite possibly superfluous.
+
+Fortunately, researchers have recently developed techniques that reduce the complexity of the genome and allow discovery and typing of thousands or tens of thousands of genome-wide makers in many individuals in a single step.
+RAD-seq is one such simple, inexpensive reduced representation technique which allows for the sequencing of small fragments of the genome adjacent to restriction enzyme cut sites (Baird et al., 2008).
+These RAD tags, or restriction-site association DNA tags, were originally developed for use in microarray hybridization typing (Miller et al., 2007), but an updated protocol substitutes second-generation DNA sequencing to rapidly discover and type SNPs (Baird et al., 2008).
+The lack of reliance on a reference genome and applicability to datasets of many individuals make it a promising technique for phylogenetic or population genetic studies in non-model organisms, such as many primates.
+
+In the present study, we summarize the RAD-seq method in brief and note the many and varied applications since its development.
+We go on to discuss reasons that make RAD-seq promising for primatologists, and then demonstrate the technique in 5 primates: a lemur, New World monkey, Old World monkey, and two apes.
+
+\subsection{The RAD-seq Technique}
+
+The following is a summary of the RAD tag library preparation of Etter et al., 2011 (\textcolor{red}{Fig.
+1}).
+The RAD-seq library preparation begins when genomic DNA from multiple individuals is digested with a restriction enzyme, such as \emph{EcoRI} or \emph{PspXI}.
+The P1 adapter is then ligated to the fragments, connected to the sticky end at the restriction enzyme cut site.
+The P1 adapter contains an amplification site for PCR, an Illumina sequencing priming site, and an individual-specific barcode of five basepairs.
+Once the barcode has been added, fragments from multiple individuals can be pooled, and the DNA is randomly sheared with a sonicator to have a length distribution under 1 kilobase.
+To select for reads that are suitable for sequencing on the Illumina platform, the sheared samples are size selected via agarose gel electrophoresis, extracting fragments between 300 and 500 bp in length.
+The second adapter, P2, is a Y adapter meaning its two halves are complementary for only part of their length.
+It is ligated to the fragments and then the fragments are amplified via PCR.
+Because the second adapter has divergent ends, the reverse amplification primer is unable to bind until after the forward amplification primer has filled in its complementary sequence.
+This ensures that only RAD tags ligated to P1 are able to amplify.
+After few cycles of PCR to minimize the risk of introducing PCR artifacts or biases, the library is ready for final clean-up, quality control, and sequencing.
+
+\subsection{Previous RAD-seq Studies}
+
+\textcolor{red}{[Previous RADseq studies. Focus on pop gen, phylogeny studies]}
+
+\subsection{Prospective Primate Applications}
+
+\textcolor{red}{[Why it would be good in primates.]}
+
+Changing the restriction enzyme to a more or less frequent cutter allows fine grained control over the number of loci, number of individuals multiplexed, and depth of coverage, allowing researchers to tailor marker discovery to their hypothesis.
 
@@
 
-	Genomic DNA from 6 primates was digested with \emph{PspXI} (New England Biolabs) and used to create a multiplexed RAD tag library.
+	We digested genomic DNA from 5 primates with \emph{PspXI} (New England Biolabs) and used it to create a multiplexed RAD tag library.
 Our library preparation method followed that of Etter et al, 2011 with the following modifications: the P1 adapter top\textcolor{red}{(?)} oligonucleotide was modified to have an overhang corresponding to the cut site of \emph{PspXI}, and a longer P2 adapter suitable for paired end sequencing was used (P2\_top: 5'-\textcolor{red}{SEQUENCEHERE}-3'; P2\_bottom: 5'-\textcolor{red}{SEQUENCEHERE}-3').
@@
 
-\paragraph{Sequence Analysis}
+\paragraph{Sequence Analysis - Clustering and SNP Discovery}
 
-	Sequence reads were demultiplexed, or separated by barcode, and reads without an expected barcode or an intact restriction enzyme cut site were excluded from the analysis.
-Reads were then aligned to the rhesus macaque reference genome (v.1.0, Mmul\_051212/rheMac2, \textcolor{red}{ref}) using BWA with default parameters.
-Reads that were unmapped, unpaired, duplicates, or that had low mapping quality were removed after alignment using Picard (\textcolor{red}{ref}) and BamTools (\textcolor{red}{ref}).
+We demultiplexed, or separated by barcode, sequence reads and excluded from the analysis reads without an expected barcode or an intact restriction enzyme cut site.
+We also removed reads with any quality scores below 10.
+Using the program Stacks, we clustered all reads into matching sets and compared closely related sets to detect orthologous loci and SNPs using a maximum likelihood approach (\textcolor{red}{Stacks ref}).
+We tallied orthologous SNPs using VCFtools (\textcolor{red}{ref}).
 
-	After performing local realignment around indels with GATK (\textcolor{red}{ref}), SNPs and short indels were identified using SAMtools mpileup and BCFtools (\textcolor{red}{ref}).
-A minimum coverage of 3 reads and a maximum of 100 was required to call a SNP or an indel at a given location.
-Orthologous SNPs were tallied using VCFtools (\textcolor{red}{ref}).
+\textcolor{red}{[Phylogeny?]}
 
-	To assess how many restriction sites were successfully sequenced and to analyze the degree of overlap between multiplexed individual's datasets, we first found all possible \emph{PspXI} cut sites using the oligoMatch utility in the USCS Genome Browser program (\textcolor{red}{ref}).
+\paragraph{Sequence Analysis - Assess RAD Tag Coverage}
+
+To assess the RAD tag coverage, we mapped human and chimpanzee reads to the human reference genome.
+Again, we excluded reads without an expected barcode or an intact restriction enzyme cut site.
+We aligned reads to the human reference genome (\textcolor{red}{hg19, ref}) using BWA with default parameters.
+We removed reads that were unmapped or that had low mapping quality after alignment using Picard (\textcolor{red}{ref}) and BamTools (\textcolor{red}{ref}).
+
+\textcolor{red}{Cut this paragraph?} After performing local realignment around indels with GATK (\textcolor{red}{ref}), we identified SNPs and short indels using SAMtools mpileup and BCFtools (\textcolor{red}{ref}).
+We required a minimum coverage of 3 reads and a maximum of 100 to call a SNP or an indel at a given location.
+We tallied orthologous SNPs using VCFtools (\textcolor{red}{ref}).
+
+To assess how many restriction sites were successfully sequenced and to analyze the degree of overlap between multiplexed individual's datasets, we first found all possible \emph{PspXI} cut sites in the human genome using the oligoMatch utility in the USCS Genome Browser program (\textcolor{red}{ref}).
 This allowed us to calculate the coverage of these restriction site-associated regions using BEDtools' multiBamCov program (\textcolor{red}{ref}).
-
-%\paragraph{Analysis Pipeline - Inferring Phylogeny}
-%\begin{itemize}
-%	\item Using Stacks? 
-%	\item Using method like cichlid people?
-%	\item Using method like Rubin et al %http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0033394
-%\end{itemize}
 
@@
 
-6.1 million sequencing reads with an intact barcode and restriction enzyme cut site could be assigned confidently to one of the six Old World monkeys in the present study.
-Roughly 2.0 million of those reads were successfully mapped to the rhesus macaque genome and passed all quality control filtration steps.
-Relative to the reference genome, our study identified 531,175 SNPs and 24,260 small indels among all samples.
+\_\_\_ million sequencing reads with an intact barcode and restriction enzyme cut site could be assigned confidently to one of the five primates in the present study.
+Roughly \_\_\_ million of those reads passed quality control filtration and \_\_\_ could be assigned to a stack.
+In these stacks our study identified \_\_\_ SNPs among all samples.
 Information for each individual is summarized in Table 1.
 
-In the rhesus macaque genome, we found 108,727 possible cut sites for \emph{PspXI}.
-
 \begin{table}[h]
-\caption{Individual Sample Information}
+\caption{Clustered Reads Data}
 \begin{center}
@@
 		\hline
-		Taxon & Source & \# Reads & \# Filtered Reads & \# Loci $\ge 1$ Read & \# Loci $\ge 3$ Reads & \# SNPs \\ \hline\hline
-		\textcolor{red}{Langur?} & \textcolor{red}{Unknown} & 1,367,050 & 306,248 & & & 363,501 \\ \hline
-		\emph{Allenopithecus nigroviridis} & \textcolor{red}{Unknown} & 1,325,558 & 447,279 & & & 356,750 \\ \hline
-		\emph{Macaca mulatta} & \textcolor{red}{Unknown} & 481,376 & 234,659 & & & 279,350 \\ \hline
-		\emph{Cercocebus torquatus atys} & \textcolor{red}{Unknown} & 297,290 & 124,700 & & & 319,766 \\ \hline
-		\emph{Papio anubis\textcolor{red}{?}} & \textcolor{red}{Unknown} & 743,556 & 313,676 & & & 316,118 \\ \hline
-		\emph{Theropithecus gelada} & \textcolor{red}{Unknown} & 1,911,030 & 603,024 & & & 366,097 \\
+		Taxon & Source & \# Reads & \# Filtered Reads & \# Stacks & Mean Coverage & \# SNPs \\ \hline\hline
+		\emph{Microcebus \textcolor{red}{murinus?}} & \textcolor{red}{Unknown} &  &  &  &  &  \\ \hline
+		\emph{Cebus \textcolor{red}{sp.?}} & \textcolor{red}{Unknown} &  &  &  &  &  \\ \hline
+		\emph{Theropithecus gelada} & \textcolor{red}{Unknown} &  &  &  &  &  \\ \hline
+		\emph{Pan troglodytes} & \textcolor{red}{Unknown} &  &  &  &  &  \\ \hline
+		\emph{Homo sapiens} & \textcolor{red}{Unknown} &  &  &  &  &  \\ \hline
 		\hline
@@
 
+In the human genome, we found \_\_\_ possible cut sites for \emph{PspXI} and \_\_\_ possible sequencing sites (two per cut site, one upstream and one downstream).
+Of those, \_\_\_ sites (\_\_\_\%) were covered by at least one read in human and \_\_\_ sites (\%) in chimpanzee.
+For \_\_\_ sites (\_\_\%), both chimpanzee and human had at least one read.
+When we restrict the analysis to sites with at least three reads, \_\_\_ sites (\_\_\_\%) were covered in human, \_\_\_ sites (\_\_\_\%) were covered in chimpanzee, and \_\_\_ sites (\_\_\_\%) in both.
+
+\begin{table}[h]
+\caption{Mapped Reads Data}
+\begin{center}
+	\small
+	\begin{tabular}{ p{3cm} || p{1.75cm} | p{1.75cm} || p{1.75cm} | p{1.75cm} | l }
+		\hline
+		Taxon & \# Reads & \# Filtered Reads & \# Loci $\ge 1$ Read & \# Loci $\ge 3$ Reads & \# SNPs \\ \hline\hline
+		\emph{Pan troglodytes} &  &  &  &  &  \\ \hline
+		\emph{Homo sapiens} &  &  &  &  &  \\ \hline
+		\hline
+	\end{tabular}
+\end{center}
+\end{table}
+
+%\begin{itemize}
+%	\item Pairwise comparisons between individuals? Like X orthologous regions shared between Papio and Thero?
+%\end{itemize}
+%
+%%\includegraphics{figs/seq_site_coverage_by_ind}
+%
+%\begin{itemize}
+%	\item SNP info from merged analysis
+%	\item SNP Venn diagram?
+%	\item Count orthologous SNPs shared between individuals. Pairwise?
+%\end{itemize}
+
+\section{Discussion}
+\textcolor{red}{
 \begin{itemize}
-	\item Number of loci hit
-	\item Number of loci hit with coverage $\ge N$
-	\item SNP info from merged analysis
-	\item SNP Venn diagram?
-	\item Calculate coverage of restriction site-associated regions
-	\item Total number possible targets in rhesus genome (compare to human too?)
-	\item Total possible target BP
-	\item How many targets did we hit?
-	\item BEDtools multiBamCoverage for this job
-	\item Number and percentage of targets with coverage $\ge 1$
-	\item Number and percentage of targets with coverage $\ge N$
-	\item Count orthologous SNPs shared between individuals
-	\item VCFtools vcf-compare for this job
+	\item RADseq is viable tool for researcher interested in primate phylogenetics, pop. gen.
+	\begin{itemize}
+		\item Cheap and easy to create libraries. Competent lab can do it in two days. (Ha!)
+		\item Multiplexing is great for pop gen studies.
+		\item Enzyme choice allows control over coverage, number of individuals, number of loci.
+		\item Baird: "different marker densities can be attained by choice of restriction enzyme"
+	\end{itemize}
+	\item Potential problems with RADseq method
+	\item ddRADseq
+	\item Promise for primatology
+\end{itemize}
+}
+
+\section{Acknowledgements}
+\begin{itemize}
+	\item NYU Med School folk
+	\item \textcolor{red}{Grants?}
 \end{itemize}
 
-\section{Discussion}
-\begin{itemize}
-	\item RADseq is viable tool for researcher interested in primate phylogenetics, pop. gen.
-	\item Enzyme choice allows control over coverage, number of individuals, number of loci.
-	\item Potential problems with RADseq method
-	\item Promise for primatology
-\end{itemize}
-
-\section{Acknowledgements}
-Acknowledgements
-
 \end{document}

### 2012-12-12 paper/paper.tex
Incorporated Luca's edits

@@
 
-During the first decades of molecular primatology, a formidable impediment to researchers was the need to develop and type polymorphic markers in a taxon of interest.
-The markers that resulted from this tedious and expensive task were often uninformative when applied outside of the population used in their design, necessitating further rounds of primer design or microsatellite assays, for example (Davey et al., 2011).
-Due to the bottleneck of marker discovery, many population genetic or phylogenetic studies in molecular primatology have been based on one or few loci (\textcolor{red}{Ting and Sterner, published yet?}).
-Such inferences can reliably give the evolutionary history of those particular regions of the genome, but they fail to adequately capture the complete complex history of the population given the mosaic nature of genomic evolution (\textcolor{red}{some gene-tree species tree review ref}).
-Adequate resolution depends on high marker density, and until recently that goal has been out of reach for many primate researchers.
+[What is mol.
+primatology?
+Explain importance of developing markers across taxa in order to study primates at different levels (population, phylogeography, phylogenetics)] During the first decades of molecular primatology, an impediment to researchers was the need to develop and type polymorphic markers in a taxon of interest.
+The markers that resulted from this time-consuming and expensive task were often uninformative when applied outside of the population used in their design, necessitating further rounds of primer design or microsatellite assays, for example (Davey et al., 2011).
+\textcolor{red}{Due to the bottleneck of marker discovery}, many population genetic or phylogenetic studies in molecular primatology have been based on one or few loci, usually mitochondrial DNA or few microsatellites (\textcolor{red}{Ting and Sterner, 2012}).
+Such inferences can reliably give the evolutionary history of those particular regions of the genome, but they fail to adequately capture the complete complex history of the population given the mosaic nature of genomic evolution (Degnan and Rosenberg, 2009; Maddison, 1997; Edwards, 2009; Maddison and Knowles, 2006).
+Adequate resolution depends on high marker density, and until recently that goal has been out of reach for many primate researchers (Edwards, 2009).
 
-The rapidly decreasing costs of DNA sequencing technology have promised revolutionary gains for primatology (\textcolor{red}{Enard and Paabo 2004, maybe?
-Goodman et al 2005?}).
+The rapidly decreasing costs of DNA sequencing technology have promised revolutionary gains for primatology (Enard and Paabo 2004; Goodman et al 2005; Ting and Sterner, 2012).
 A primate researcher benefits from the many nearby sequenced and assembled reference genomes in the order, but genomic studies of non-model organisms nevertheless remain difficult.
-Though the cost of low coverage whole genome sequencing has fallen to a level feasible for many researchers' budgets, sequencing whole genomes for the tens or hundreds of individuals desired in a typical population genetic study is often prohibitively expensive and quite possibly superfluous.
-
-Fortunately, researchers have recently developed techniques that reduce the complexity of the genome and allow discovery and typing of thousands or tens of thousands of genome-wide makers in many individuals in a single step.
+Though the cost of whole genome sequencing has fallen to a level feasible for many researchers' budgets (e.g., Perry et al., 2012), sequencing whole genomes for the tens or hundreds of individuals desired in a typical population genetic study is often prohibitively expensive and quite possibly superfluous (McCormack, et al.
+2012).
+Fortunately, researchers have recently developed techniques that reduce the complexity of the genome and allow discovery and typing of thousands or tens of thousands of genome-wide makers in many individuals in a single step (Davey et al., 2011; McCormack, et al.
+2012).
 RAD-seq is one such simple, inexpensive reduced representation technique which allows for the sequencing of small fragments of the genome adjacent to restriction enzyme cut sites (Baird et al., 2008).
-These RAD tags, or restriction-site association DNA tags, were originally developed for use in microarray hybridization typing (Miller et al., 2007), but an updated protocol substitutes second-generation DNA sequencing to rapidly discover and type SNPs (Baird et al., 2008).
+These RAD tags, or restriction-site association DNA tags, were originally developed for use in microarray hybridization typing (Miller et al., 2007), but an updated protocol substitutes second-generation DNA sequencing to rapidly discover and type SNPs (Baird et al., 2008; Etter et al., 2011).
 The lack of reliance on a reference genome and applicability to datasets of many individuals make it a promising technique for phylogenetic or population genetic studies in non-model organisms, such as many primates.
@@
 1}).
-The RAD-seq library preparation begins when genomic DNA from multiple individuals is digested with a restriction enzyme, such as \emph{EcoRI} or \emph{PspXI}.
+The RAD-seq library preparation begins when genomic DNA from multiple individuals is digested with a restriction enzyme, such as \emph{EcoRI} or \emph{PspXI} (1a).
 The P1 adapter is then ligated to the fragments, connected to the sticky end at the restriction enzyme cut site.
-The P1 adapter contains an amplification site for PCR, an Illumina sequencing priming site, and an individual-specific barcode of five basepairs.
-Once the barcode has been added, fragments from multiple individuals can be pooled, and the DNA is randomly sheared with a sonicator to have a length distribution under 1 kilobase.
+The P1 adapter contains an amplification site for PCR, an Illumina sequencing priming site, and an individual-specific barcode of five basepairs (1b).
+Once the barcode has been added, fragments from multiple individuals can be pooled, and the DNA is randomly sheared with a sonicator to have a length distribution under 1 kilobase (1c).
 To select for reads that are suitable for sequencing on the Illumina platform, the sheared samples are size selected via agarose gel electrophoresis, extracting fragments between 300 and 500 bp in length.
-The second adapter, P2, is a Y adapter meaning its two halves are complementary for only part of their length.
-It is ligated to the fragments and then the fragments are amplified via PCR.
+The second adapter, P2, is a Y adapter meaning its two halves are complementary for only part of their length (1d).
+It is ligated to the fragments and then the fragments are amplified via PCR (1e).
 Because the second adapter has divergent ends, the reverse amplification primer is unable to bind until after the forward amplification primer has filled in its complementary sequence.
@@
 
-\textcolor{red}{[Previous RADseq studies. Focus on pop gen, phylogeny studies]}
+Restriction site-associated DNA Sequencing (RAD-Seq) is an economical and efficient method for SNP discovery and genotyping.
+Since its first application by Baird and colleagues (2008) on two model organisms – the fungus \emph{Neurospora crassa} and the three-spined stickleback, \emph{Gasterosteus aculeatus} – RAD-seq has been successfully applied to several organisms for which reference genome information was not available.
 
-\subsection{Prospective Primate Applications}
+The ability of RAD-seq technology to identify thousands of orthologous SNPs across multiple individuals at both intra- and interspecific level makes this technique extremely promising for the study of population structure (Hohenlohe et al., 2010; Emerson et al., 2010; Keller et al., 2012), gene flow and hybridization (Hohenlohe et al., 2011; Keller et al., 2012), phylogeography (Emerson et al., 2010) and phylogeny (Wagner et al., 2012; Rubin et al., 2012).
+The RAD-tag sequencing approach has been particularly used to generate SNP data to address questions in population genomics.
+For example, a series of studies conducted by Hohenlohe and colleagues investigated parallel adaptation and hybridization in several species of fish (Hohenlohe et al.
+2010, 2011, 2012), while Emerson et al (2010) identified more than 3,700 SNPs for pitcher plant mosquitoes in eastern North America to provide the first phylogeographic study using RAD sequence data.
 
-\textcolor{red}{[Why it would be good in primates.]}
-
-Changing the restriction enzyme to a more or less frequent cutter allows fine grained control over the number of loci, number of individuals multiplexed, and depth of coverage, allowing researchers to tailor marker discovery to their hypothesis.
+Although RAD sequencing is more effective in addressing questions at or below the level of a single species, a few recent studies have used this technique in the analysis of phylogenetic questions.
+Rubin et al (2012) provided a simulation study in which they investigated the accuracy of RAD-seq data to reconstruct phylogenies in organisms with different population sizes and clade ages (\emph{Drosophila}, mammals, and yeasts).
+In their study the authors supported the efficiency of RAD-seq data in inferring phylogenies, but they also caution about how this approach achieves the best results in younger clades where more orthologous restriction sites are likely to be retained across species.
+This simulation analysis was confirmed in two recent empirical studies in which a RAD-tag sequencing approach was successfully used to reconstruct phylogenetic relationships in two recent but speciose radiation, such as the African cyclids (Wagner et al., 2012) and the \emph{Heliconius} butterﬂies (Nadeau et al., 2012).
 
@@
 \section{Discussion}
-\textcolor{red}{
-\begin{itemize}
-	\item RADseq is viable tool for researcher interested in primate phylogenetics, pop. gen.
+
+\textcolor{red}{[Why it would be good in primates.]}
+
+RADseq is viable tool for researcher interested in primate phylogenetics, pop. gen.
+
 \begin{itemize}
@@
 \end{itemize}
-	\item Potential problems with RADseq method
-	\item ddRADseq
-	\item Promise for primatology
-\end{itemize}
-}
+
+Changing the restriction enzyme to a more or less frequent cutter allows fine grained control over the number of loci, number of individuals multiplexed, and depth of coverage, allowing researchers to tailor marker discovery to their hypothesis.
+
+Potential problems with RADseq method including the phylogenetic distance and the need of high quality DNA (at least for now)
+
+ddRADseq
 

### 2012-12-12 trunk/paper/paper.tex
Finished clustering part of results section

--- 2012-12-09 trunk/paper/paper.tex
@@
 \section{Introduction}
-\begin{itemize}
-	\item Next-gen sequencing revolution promises gains in primatology
-	\item Still expensive
-	\item Many genomes, but still tough doing genomics on non-model organisms
-	\item What is RADseq?
-	\item Previous RADseq studies
-	\item Why would it be good for primates
-	\item PRESENT STUDY
-	\begin{itemize}
-		\item We did RADseq on 6 Cercopithecoids
-		\item Assessed how well it worked
-		\item Show it has promise for primates
-	\end{itemize}
-\end{itemize}
+
+\textcolor{red}{[What is mol.
+primatology?
+Explain importance of developing markers across taxa in order to study primates at different levels (population, phylogeography, phylogenetics)]} During the first decades of molecular primatology, an impediment to researchers was the need to develop and type polymorphic markers in a taxon of interest.
+The markers that resulted from this time-consuming and expensive task were often uninformative when applied outside of the population used in their design, necessitating further rounds of primer design or microsatellite assays, for example (Davey et al., 2011).
+Due to the bottleneck caused by the inefficiency of marker discovery, many population genetic or phylogenetic studies in molecular primatology have been based on one or few loci, usually mitochondrial DNA or few microsatellites (Ting and Sterner, 2012).
+Such inferences can reliably give the evolutionary history of those particular regions of the genome, but they fail to adequately capture the complete complex history of the population given the mosaic nature of genomic evolution (Degnan and Rosenberg, 2009; Maddison, 1997; Edwards, 2009; Maddison and Knowles, 2006).
+Adequate resolution depends on high marker density, and until recently that goal has been out of reach for many primate researchers (Edwards, 2009).
+
+The rapidly decreasing costs of DNA sequencing technology have promised revolutionary gains for primatology (Enard and Paabo 2004; Goodman et al 2005; Ting and Sterner, 2012).
+A primate researcher benefits from the many nearby sequenced and assembled reference genomes in the order, but genomic studies of non-model organisms nevertheless remain difficult.
+Though the cost of whole genome sequencing has fallen to a level feasible for many researchers' budgets (e.g., Perry et al., 2012), sequencing whole genomes for the tens or hundreds of individuals desired in a typical population genetic study is often prohibitively expensive and quite possibly superfluous (McCormack, et al.
+2012).
+Fortunately, researchers have recently developed techniques that reduce the complexity of the genome and allow discovery and typing of thousands or tens of thousands of genome-wide makers in many individuals in a single step (Davey et al., 2011; McCormack, et al.
+2012).
+RAD-seq is one such simple, inexpensive reduced representation technique which allows for the sequencing of small fragments of the genome adjacent to restriction enzyme cut sites (Baird et al., 2008).
+These RAD tags, or restriction-site association DNA tags, were originally developed for use in microarray hybridization typing (Miller et al., 2007), but an updated protocol substitutes second-generation DNA sequencing to rapidly discover and type SNPs (Baird et al., 2008; Etter et al., 2011).
+The lack of reliance on a reference genome and applicability to datasets of many individuals make it a promising technique for phylogenetic or population genetic studies in non-model organisms, such as many primates.
+
+In the present study, we summarize the RAD-seq method in brief and note the many and varied applications since its development.
+We go on to discuss reasons that make RAD-seq promising for primatologists, and then demonstrate the technique in 5 primates: a lemur, New World monkey, Old World monkey, and two apes.
+
+\subsection{The RAD-seq Technique}
+
+The following is a summary of the RAD tag library preparation of Etter et al., 2011 (\textcolor{red}{Fig.
+1}).
+The RAD-seq library preparation begins when genomic DNA from multiple individuals is digested with a restriction enzyme, such as \emph{EcoRI} or \emph{PspXI} (1a).
+The P1 adapter is then ligated to the fragments, connected to the sticky end at the restriction enzyme cut site.
+The P1 adapter contains an amplification site for PCR, an Illumina sequencing priming site, and an individual-specific barcode of five basepairs (1b).
+Once the barcode has been added, fragments from multiple individuals can be pooled, and the DNA is randomly sheared with a sonicator to have a length distribution under 1 kilobase (1c).
+To select for reads that are suitable for sequencing on the Illumina platform, the sheared samples are size selected via agarose gel electrophoresis, extracting fragments between 300 and 500 bp in length.
+The second adapter, P2, is a Y adapter meaning its two halves are complementary for only part of their length (1d).
+It is ligated to the fragments and then the fragments are amplified via PCR (1e).
+Because the second adapter has divergent ends, the reverse amplification primer is unable to bind until after the forward amplification primer has filled in its complementary sequence.
+This ensures that only RAD tags ligated to P1 are able to amplify.
+After few cycles of PCR to minimize the risk of introducing PCR artifacts or biases, the library is ready for final clean-up, quality control, and sequencing.
+
+\subsection{Previous RAD-seq Studies}
+
+Restriction site-associated DNA Sequencing (RAD-Seq) is an economical and efficient method for SNP discovery and genotyping.
+Since its first application by Baird and colleagues (2008) on two model organisms – the fungus \emph{Neurospora crassa} and the three-spined stickleback, \emph{Gasterosteus aculeatus} – RAD-seq has been successfully applied to several organisms for which reference genome information was not available.
+
+The ability of RAD-seq technology to identify thousands of orthologous SNPs across multiple individuals at both intra- and interspecific level makes this technique extremely promising for the study of population structure (Hohenlohe et al., 2010; Emerson et al., 2010; Keller et al., 2012), gene flow and hybridization (Hohenlohe et al., 2011; Keller et al., 2012), phylogeography (Emerson et al., 2010) and phylogeny (Wagner et al., 2012; Rubin et al., 2012).
+The RAD-tag sequencing approach has been particularly used to generate SNP data to address questions in population genomics.
+For example, a series of studies conducted by Hohenlohe and colleagues investigated parallel adaptation and hybridization in several species of fish (Hohenlohe et al.
+2010, 2011, 2012), while Emerson et al (2010) identified more than 3,700 SNPs for pitcher plant mosquitoes in eastern North America to provide the first phylogeographic study using RAD sequence data.
+
+Although RAD sequencing is more effective in addressing questions at or below the level of a single species, a few recent studies have used this technique in the analysis of phylogenetic questions.
+Rubin et al (2012) provided a simulation study in which they investigated the accuracy of RAD-seq data to reconstruct phylogenies in organisms with different population sizes and clade ages (\emph{Drosophila}, mammals, and yeasts).
+In their study the authors supported the efficiency of RAD-seq data in inferring phylogenies, but they also caution about how this approach achieves the best results in younger clades where more orthologous restriction sites are likely to be retained across species.
+This simulation analysis was confirmed in two recent empirical studies in which a RAD-tag sequencing approach was successfully used to reconstruct phylogenetic relationships in two recent but speciose radiation, such as the African cyclids (Wagner et al., 2012) and the \emph{Heliconius} butterﬂies (Nadeau et al., 2012).
 
@@
 
-	Genomic DNA from 6 animals was digested with \emph{PspXI} (New England Biolabs) and used to create a multiplexed RAD tag library.
+	We digested genomic DNA from 5 primates with \emph{PspXI} (New England Biolabs) and used it to create a multiplexed RAD tag library.
 Our library preparation method followed that of Etter et al, 2011 with the following modifications: the P1 adapter top\textcolor{red}{(?)} oligonucleotide was modified to have an overhang corresponding to the cut site of \emph{PspXI}, and a longer P2 adapter suitable for paired end sequencing was used (P2\_top: 5'-\textcolor{red}{SEQUENCEHERE}-3'; P2\_bottom: 5'-\textcolor{red}{SEQUENCEHERE}-3').
@@
 We sequenced the prepared library as one 150-cycle paired-end run of an Illumina MiSeq at the NYU Langone Medical Center's Genome Technology Center using a spike-in of 30\% PhiX DNA to control for low diversity in the library at the barcode and restriction sites.
+Other individuals were sequenced alongside those of the present study.
 Sequences are available to download from the NCBI Short Read Archive (accession number \textcolor{red}{SRAXXXXXX.X}).
 
+\paragraph{Sequence Analysis - Clustering and SNP Discovery}
+
+We demultiplexed, or separated by barcode, sequence reads and excluded from the analysis reads without an expected barcode or an intact restriction enzyme cut site.
+We also removed reads with any quality scores below 10.
+Using the program Stacks, we clustered all reads into sets that differed by no more than two basepairs and compared closely related sets to detect orthologous loci and SNPs using a maximum likelihood approach (\textcolor{red}{Stacks ref}).
+We tallied orthologous SNPs using VCFtools (\textcolor{red}{ref}).
+
+\paragraph{Sequence Analysis - Assess RAD Tag Coverage}
+
+To assess the RAD tag coverage, we mapped human and chimpanzee reads to the highest quality primate reference genome, that of humans.
+Again, we excluded reads without an expected barcode or an intact restriction enzyme cut site.
+We aligned reads to the human reference genome (\textcolor{red}{GRCh37/hg19, ref}) using BWA with default parameters.
+We removed reads that were unmapped or that had low mapping quality after alignment using Picard (\textcolor{red}{ref}) and BamTools (\textcolor{red}{ref}).
+
+After performing local realignment around indels with GATK (\textcolor{red}{ref}), we identified SNPs and short indels using SAMtools mpileup and BCFtools (\textcolor{red}{ref}).
+We required a minimum coverage of 3 reads and a maximum of 100 to call a SNP or an indel at a given location.
+We tallied orthologous SNPs using VCFtools (\textcolor{red}{ref}).
+
+To assess how many restriction sites were successfully sequenced and to analyze the degree of overlap between multiplexed individual's datasets, we first found all possible \emph{PspXI} cut sites in the human genome using the oligoMatch utility in the USCS Genome Browser program (\textcolor{red}{ref}).
+This allowed us to calculate the coverage of these restriction site-associated regions using BEDtools' multiBamCov program (\textcolor{red}{ref}).
+
+\section{Results}
+
+12.3 million sequencing reads with an intact barcode and restriction enzyme cut site could be assigned confidently to one of the five primates in the present study.
+Roughly 9.1 million of those reads passed quality control filtration and were clustered into stacks.
+By comparing these stacks and including only SNPs that were present in multiple individuals, our study identified 7,910 SNPs among all samples.
+Information for each individual is summarized in Table 1.
+
 \begin{table}[h]
-\caption{Samples}
+\caption{Clustered Reads Data}
 \begin{center}
-  \begin{tabular}{ l | l | l }
+	\small
+	\begin{tabular}{ p{3cm} || l || p{1.75cm} | p{1.75cm} || p{1.75cm} | p{1.75cm} | l }
 		\hline
-    Taxon & Source & Geographic Locality \\ \hline\hline
-    \textcolor{red}{Langur?} & NYU Collection & Unknown \\ \hline
-    \emph{Allenopithecus nigroviridis} & NYU Collection & Unknown \\ \hline
-    \emph{Macaca mulatta} & NYU Collection & Unknown \\ \hline
-    \emph{Cercocebus torquatus atys} & NYU Collection & Unknown \\ \hline
-    \emph{Papio anubis\textcolor{red}{?}} & NYU Collection & Unknown \\ \hline
-    \emph{Theropithecus gelada} & NYU Collection & Unknown \\ 
-    \hline
+		Taxon                                  & Source                   & \# Reads & \# Filtered Reads & \# Ind. Stacks & Mean Coverage (SD)  & \# Shared SNPs \\ \hline\hline
+		\emph{Microcebus \textcolor{red}{sp?}} & \textcolor{red}{Unknown} &  2,830,832 & 2,025,103       & 248,324         &  7.66 (20.71)  & 13 \\ \hline
+		\emph{Cebus \textcolor{red}{sp.?}}     & \textcolor{red}{Unknown} &  1,946,096 & 1,427,413       & 107,829         & 12.64 (139.08) & 56 \\ \hline
+		\emph{Theropithecus gelada}            & \textcolor{red}{Unknown} &  1,918,425 & 1,392,709       & 136,657         &  9.70 (17.34)  & 212 \\ \hline
+		\emph{Pan troglodytes}                 & \textcolor{red}{Unknown} &  2,616,062 & 1,910,560       & 157,775         & 11.50 (42.27)  & 5,886 \\ \hline
+		\emph{Homo sapiens}                    & \textcolor{red}{Unknown} &  3,032,823 & 2,374,733       & 131,544         & 17.37 (25.30)  & 5,786 \\ \hline
 	\end{tabular}
@@
 
-\paragraph{Sequence Analysis}
+In the human genome, we found \_\_\_ possible cut sites for \emph{PspXI} and \_\_\_ possible sequencing sites (two per cut site, one upstream and one downstream).
+Of those, \_\_\_ sites (\_\_\_\%) were covered by at least one read in human and \_\_\_ sites (\%) in chimpanzee.
+For \_\_\_ sites (\_\_\%), both chimpanzee and human had at least one read.
+When we restrict the analysis to sites with at least three reads, \_\_\_ sites (\_\_\_\%) were covered in human, \_\_\_ sites (\_\_\_\%) were covered in chimpanzee, and \_\_\_ sites (\_\_\_\%) in both.
 
-	Sequence reads were demultiplexed, or separated by barcode, and reads without an expected barcode or an intact restriction enzyme cut site were excluded from the analysis.
-Reads were then aligned to the rhesus macaque reference genome (v.1.0, Mmul\_051212/rheMac2, \textcolor{red}{ref}) using BWA with default parameters.
-Reads that were unmapped, unpaired, duplicates, or that had low mapping quality were removed after alignment using Picard (\textcolor{red}{ref}) and BamTools (\textcolor{red}{ref}).
+\begin{table}[h]
+\caption{Mapped Reads Data}
+\begin{center}
+	\small
+	\begin{tabular}{ p{3cm} || p{1.75cm} | p{1.75cm} || p{1.75cm} | p{1.75cm} | l }
+		\hline
+		Taxon & \# Reads & \# Filtered Reads & \# Loci $\ge 1$ Read & \# Loci $\ge 3$ Reads & \# SNPs \\ \hline\hline
+		\emph{Pan troglodytes} & 2,616,062 &  &  &  &  \\ \hline
+		\emph{Homo sapiens}    & 3,032,823 &  &  &  &  \\ \hline
+	\end{tabular}
+\end{center}
+\end{table}
 
-	After performing local realignment around indels with GATK (\textcolor{red}{ref}), SNPs and short indels were identified using SAMtools mpileup and BCFtools (\textcolor{red}{ref}).
-A minimum coverage of 3 reads and a maximum of 100 was required to call a SNP or an indel at a given location.
+\section{Discussion}
 
-	To analyze the degree of overlap between multiplexed individual's datasets, we...
+\textcolor{red}{[Why it would be good in primates.]}
 
-\paragraph{Analysis Pipeline - Analysis of Degree of Overlap}
+RADseq is viable tool for researcher interested in primate phylogenetics, pop. gen.
+
 \begin{itemize}
-	\item Calculate coverage of restriction site-associated regions
-	\begin{itemize}
-		\item Info on targeted intervals
-		\begin{itemize}
-			\item Total number possible targets in rhesus genome (compare to human too?)
-			\item Total possible target BP
-		\end{itemize}
-		\item How many targets did we hit?
-		\begin{itemize}
-			\item BEDtools multiBamCoverage for this job
-			\item Number and percentage of targets with coverage $\ge 1$
-			\item Number and percentage of targets with coverage $\ge N$
-		\end{itemize}
-	\end{itemize}
-	\item Count orthologous SNPs shared between individuals
-	\begin{itemize}
-		\item VCFtools vcf-compare for this job
-	\end{itemize}
+	\item Cheap and easy to create libraries. Competent lab can do it in two days. (Ha!)
+	\item Multiplexing is great for pop gen studies.
+	\item Enzyme choice allows control over coverage, number of individuals, number of loci.
+	\item Baird: "different marker densities can be attained by choice of restriction enzyme"
 \end{itemize}
 
-%\paragraph{Analysis Pipeline - Inferring Phylogeny}
-%\begin{itemize}
-%	\item Using Stacks? 
-%	\item Using method like cichlid people?
-%	\item Using method like Rubin et al %http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0033394
-%\end{itemize}
+Changing the restriction enzyme to a more or less frequent cutter allows fine grained control over the number of loci, number of individuals multiplexed, and depth of coverage, allowing researchers to tailor marker discovery to their hypothesis.
 
-\section{Results}
+Potential problems with RADseq method including the phylogenetic distance and the need of high quality DNA (at least for now)
+
+RADseq is shitty when phylogenetic distances are great.
+Works much better within a genus or species level.
+We managed to infer the accepted primate phylogeny from our SNP data using maximum likelihood approach, but the support values were crap
+
+ddRADseq
+
+\section{Acknowledgements}
 \begin{itemize}
-	\item Info from analyzing reads with FastQC
-	\begin{itemize}
-		\item Number of reads (per ind. too see table)
-		\item Total sequence bp (per ind. too see table)
-		\item Maximum possible sequence depth (Cut?)
-		\item Other stats that FastQC gives you?
-	\end{itemize}
-	\item Table: 
-	\begin{itemize}
-		\item Number of reads per animal
-		\item Total sequenced bp per animal?
-		\item Number that passed filtration
-		\item Number of loci hit
-		\item Number of loci hit with coverage $\ge N$
-		\item Number of SNPs
-	\end{itemize}
-	\item SNP info from merged analysis
-	\item SNP Venn diagram?
-	\item Table of overlapping region, orthologous SNP counts
-	%\item Phylogenetic tree
+	\item NYU Med School folk
+	\item \textcolor{red}{Grants?}
 \end{itemize}
 
-\section{Discussion}
-\begin{itemize}
-	\item RADseq is viable tool for researcher interested in primate phylogenetics, pop. gen.
-	\item Enzyme choice allows control over coverage, number of individuals, number of loci.
-	\item Potential problems with RADseq method
-	\item Promise for primatology
-\end{itemize}
-
-\section{Acknowledgements}
-Acknowledgements
-
 \end{document}

### 2012-12-12 trunk/paper/paper.tex
Added count of PspXI cut sites

@@
 
-To assess how many restriction sites were successfully sequenced and to analyze the degree of overlap between multiplexed individual's datasets, we first found all possible \emph{PspXI} cut sites in the human genome using the oligoMatch utility in the USCS Genome Browser program (\textcolor{red}{ref}).
+To assess how many restriction sites were successfully sequenced and to analyze the degree of overlap between multiplexed individual's datasets, we first found all possible \emph{PspXI} cut sites in the human genome using the oligoMatch utility in the USCS Genome Browser program and created a BED file of all regions 1000 upstream and downstream.
+(\textcolor{red}{ref}).
 This allowed us to calculate the coverage of these restriction site-associated regions using BEDtools' multiBamCov program (\textcolor{red}{ref}).
@@
 
-In the human genome, we found \_\_\_ possible cut sites for \emph{PspXI} and \_\_\_ possible sequencing sites (two per cut site, one upstream and one downstream).
+In the human genome, we found 58,172 possible cut sites for \emph{PspXI} and 116,344 possible sequencing sites (two per cut site, one upstream and one downstream).
 Of those, \_\_\_ sites (\_\_\_\%) were covered by at least one read in human and \_\_\_ sites (\%) in chimpanzee.
@@
 \section{Acknowledgements}
+
+The authors would like to thank the NYU Langone Medical Center's Genome Technology Center for assistance with library preparation and sequencing.
+
 \begin{itemize}

### 2012-12-12 trunk/paper/paper.tex
Fixed introduction

@@
-\title{RADseq Works in Primates, Dammit.}
+\title{A new method for genome-wide marker development and typing holds great promise for primate population genetics}
 \author{
@@
 
-\textcolor{red}{[What is mol.
-primatology?
-Explain importance of developing markers across taxa in order to study primates at different levels (population, phylogeography, phylogenetics)]} During the first decades of molecular primatology, an impediment to researchers was the need to develop and type polymorphic markers in a taxon of interest.
+In molecular primatology, as in all fields of molecular biology, the development of variable genetic markers is essential for the study of organisms at different levels, from population genetic to phylogeographic to phylogenetic research.
+During the first decades of the discipline, an impediment to researchers was the need to develop and type polymorphic markers in a taxon of interest.
 The markers that resulted from this time-consuming and expensive task were often uninformative when applied outside of the population used in their design, necessitating further rounds of primer design or microsatellite assays, for example (Davey et al., 2011).
-Due to the bottleneck caused by the inefficiency of marker discovery, many population genetic or phylogenetic studies in molecular primatology have been based on one or few loci, usually mitochondrial DNA or few microsatellites (Ting and Sterner, 2012).
+Due to the bottleneck caused by the inefficiency of marker discovery, many population genetic or phylogenetic studies in molecular primatology have been based on one or few loci, usually mitochondrial DNA or a few microsatellites (Ting and Sterner, 2012).
 Such inferences can reliably give the evolutionary history of those particular regions of the genome, but they fail to adequately capture the complete complex history of the population given the mosaic nature of genomic evolution (Degnan and Rosenberg, 2009; Maddison, 1997; Edwards, 2009; Maddison and Knowles, 2006).
@@
 In the present study, we summarize the RAD-seq method in brief and note the many and varied applications since its development.
-We go on to discuss reasons that make RAD-seq promising for primatologists, and then demonstrate the technique in 5 primates: a lemur, New World monkey, Old World monkey, and two apes.
+We demonstrate the technique in 5 primates: a lemur, New World monkey, Old World monkey, and two apes, and discuss the features that make RAD-seq promising for primatologists.
 
@@
 
-The following is a summary of the RAD tag library preparation of Etter et al., 2011 (\textcolor{red}{Fig.
+The following is a summary of the RAD tag library preparation protocol of Etter et al., 2011 (\textcolor{red}{Fig.
 1}).
-The RAD-seq library preparation begins when genomic DNA from multiple individuals is digested with a restriction enzyme, such as \emph{EcoRI} or \emph{PspXI} (1a).
+The RAD-seq library preparation begins when genomic DNA is digested with a restriction enzyme, such as \emph{EcoRI} or \emph{PspXI} (1a).
 The P1 adapter is then ligated to the fragments, connected to the sticky end at the restriction enzyme cut site.

### 2012-12-13 trunk/paper/paper.tex
Done except for Luca's section at end and abstract, essentially

@@
 
-The following is a summary of the RAD tag library preparation protocol of Etter et al., 2011 (\textcolor{red}{Fig.
-1}).
-The RAD-seq library preparation begins when genomic DNA is digested with a restriction enzyme, such as \emph{EcoRI} or \emph{PspXI} (1a).
+The following is a summary of the RAD tag library preparation protocol of Etter et al., 2011 (Fig.
+1).
+The RAD-seq library preparation begins when genomic DNA is digested with a restriction enzyme, such as \emph{EcoRI} or \emph{PspXI} (1A).
 The P1 adapter is then ligated to the fragments, connected to the sticky end at the restriction enzyme cut site.
-The P1 adapter contains an amplification site for PCR, an Illumina sequencing priming site, and an individual-specific barcode of five basepairs (1b).
-Once the barcode has been added, fragments from multiple individuals can be pooled, and the DNA is randomly sheared with a sonicator to have a length distribution under 1 kilobase (1c).
+The P1 adapter contains an amplification site for PCR, an Illumina sequencing priming site, and an individual-specific barcode of five basepairs (1B).
+Once the barcode has been added, fragments from multiple individuals can be pooled (1C), and the DNA is randomly sheared with a sonicator to have a length distribution under 1 kilobase (1D).
 To select for reads that are suitable for sequencing on the Illumina platform, the sheared samples are size selected via agarose gel electrophoresis, extracting fragments between 300 and 500 bp in length.
-The second adapter, P2, is a Y adapter meaning its two halves are complementary for only part of their length (1d).
-It is ligated to the fragments and then the fragments are amplified via PCR (1e).
+The second adapter, P2, is a Y adapter meaning its two halves are complementary for only part of their length (1E).
+It is ligated to the fragments and then the fragments are amplified via PCR (1F).
 Because the second adapter has divergent ends, the reverse amplification primer is unable to bind until after the forward amplification primer has filled in its complementary sequence.
@@
 Individual-specific barcodes contained in the P1 adapter differed by at least \textcolor{red}{three} nucleotides.
-We chose \emph{PspXI} based on the results of \emph{in silico} digestion of the human, rhesus macaque, and baboon reference genomes using custom Perl scripts (\textcolor{red}{refs}).
-We sequenced the prepared library as one 150-cycle paired-end run of an Illumina MiSeq at the NYU Langone Medical Center's Genome Technology Center using a spike-in of 30\% PhiX DNA to control for low diversity in the library at the barcode and restriction sites.
+We chose \emph{PspXI} based on the results of \emph{in silico} digestion of the human, rhesus macaque, and baboon reference genomes using custom Perl scripts.
+We sequenced the prepared library as one 150-cycle paired-end run and one 150-cycle single-end run of an Illumina MiSeq at the NYU Langone Medical Center's Genome Technology Center using a spike-in of 30\% PhiX DNA to control for low diversity in the library at the barcode and restriction sites.
 Other individuals were sequenced alongside those of the present study.
@@
 
+As input for the clustering analysis, we combined the first read of the paired-end run and the single-end run reads.
 We demultiplexed, or separated by barcode, sequence reads and excluded from the analysis reads without an expected barcode or an intact restriction enzyme cut site.
 We also removed reads with any quality scores below 10.
-Using the program Stacks, we clustered all reads into sets that differed by no more than two basepairs and compared closely related sets to detect orthologous loci and SNPs using a maximum likelihood approach (\textcolor{red}{Stacks ref}).
-We tallied orthologous SNPs using VCFtools (\textcolor{red}{ref}).
+Using the program Stacks, we clustered all reads into sets that differed by no more than two basepairs and compared closely related sets to detect orthologous loci and SNPs using a maximum likelihood approach (Catchen et al., 2011).
+We tallied orthologous SNPs using VCFtools (Danecek et al., 2011).
 
@@
 Again, we excluded reads without an expected barcode or an intact restriction enzyme cut site.
-We aligned reads to the human reference genome (\textcolor{red}{GRCh37/hg19, ref}) using BWA with default parameters.
-We removed reads that were unmapped or that had low mapping quality after alignment using Picard (\textcolor{red}{ref}) and BamTools (\textcolor{red}{ref}).
+We aligned reads to the human reference genome (GRCh37/hg19, Human Genome Consortium 2001) using BWA with default parameters (Li and Durbin, 2009).
+We separately mapped the single-end and paired-end data and then combined the resultant files after alignment.
+We removed reads that were unmapped or that had low mapping quality using Picard (http://picard.sourceforge.net) and BamTools (Barnett et al., 2011).
 
-After performing local realignment around indels with GATK (\textcolor{red}{ref}), we identified SNPs and short indels using SAMtools mpileup and BCFtools (\textcolor{red}{ref}).
+After performing local realignment around indels with GATK (DePristo et al., 2011), we identified SNPs and short indels using SAMtools mpileup and BCFtools (Li et al., 2009).
 We required a minimum coverage of 3 reads and a maximum of 100 to call a SNP or an indel at a given location.
-We tallied orthologous SNPs using VCFtools (\textcolor{red}{ref}).
+We tallied orthologous SNPs using VCFtools (Danecek et al., 2011).
 
-To assess how many restriction sites were successfully sequenced and to analyze the degree of overlap between multiplexed individual's datasets, we first found all possible \emph{PspXI} cut sites in the human genome using the oligoMatch utility in the USCS Genome Browser program and created a BED file of all regions 1000 upstream and downstream.
-(\textcolor{red}{ref}).
-This allowed us to calculate the coverage of these restriction site-associated regions using BEDtools' multiBamCov program (\textcolor{red}{ref}).
+To assess how many restriction sites were successfully sequenced and to analyze the degree of overlap between multiplexed individual's datasets, we first found all possible \emph{PspXI} cut sites in the human genome using the oligoMatch utility in the USCS Genome Browser program and created a BED file of all regions 1000 basepairs upstream and downstream.
+(Meyer et al., 2012).
+This allowed us to calculate the coverage of these restriction site-associated regions using BEDtools' multiBamCov program (Quinlan and Hall 2010).
 
@@
 		\hline
-		Taxon                                  & Source                   & \# Reads & \# Filtered Reads & \# Ind. Stacks & Mean Coverage (SD)  & \# Shared SNPs \\ \hline\hline
-		\emph{Microcebus \textcolor{red}{sp?}} & \textcolor{red}{Unknown} &  2,830,832 & 2,025,103       & 248,324         &  7.66 (20.71)  & 13 \\ \hline
-		\emph{Cebus \textcolor{red}{sp.?}}     & \textcolor{red}{Unknown} &  1,946,096 & 1,427,413       & 107,829         & 12.64 (139.08) & 56 \\ \hline
-		\emph{Theropithecus gelada}            & \textcolor{red}{Unknown} &  1,918,425 & 1,392,709       & 136,657         &  9.70 (17.34)  & 212 \\ \hline
-		\emph{Pan troglodytes}                 & \textcolor{red}{Unknown} &  2,616,062 & 1,910,560       & 157,775         & 11.50 (42.27)  & 5,886 \\ \hline
-		\emph{Homo sapiens}                    & \textcolor{red}{Unknown} &  3,032,823 & 2,374,733       & 131,544         & 17.37 (25.30)  & 5,786 \\ \hline
+		Taxon                       & \# Reads  & \# Filtered Reads & \# Ind. Stacks & Mean Coverage (SD)  & \# Shared SNPs \\ \hline\hline
+		\emph{Microcebus sp.}       & 2,830,832 & 2,025,103       & 248,324         &  7.66 (20.71)  & 13    \\ \hline
+		\emph{Cebus sp.}            & 1,946,096 & 1,427,413       & 107,829         & 12.64 (139.08) & 56    \\ \hline
+		\emph{Theropithecus gelada} & 1,918,425 & 1,392,709       & 136,657         &  9.70 (17.34)  & 212   \\ \hline
+		\emph{Pan troglodytes}      & 2,616,062 & 1,910,560       & 157,775         & 11.50 (42.27)  & 5,886 \\ \hline
+		\emph{Homo sapiens}         & 3,032,823 & 2,374,733       & 131,544         & 17.37 (25.30)  & 5,786 \\ \hline
 	\end{tabular}
@@
 In the human genome, we found 58,172 possible cut sites for \emph{PspXI} and 116,344 possible sequencing sites (two per cut site, one upstream and one downstream).
-Of those, \_\_\_ sites (\_\_\_\%) were covered by at least one read in human and \_\_\_ sites (\%) in chimpanzee.
-For \_\_\_ sites (\_\_\%), both chimpanzee and human had at least one read.
-When we restrict the analysis to sites with at least three reads, \_\_\_ sites (\_\_\_\%) were covered in human, \_\_\_ sites (\_\_\_\%) were covered in chimpanzee, and \_\_\_ sites (\_\_\_\%) in both.
+Of those possible location, 111,686 sites (96.00\%) had at least one mapped read present in human and 91,646 sites (78.77\%) in chimpanzee.
+For 90,022 sites (77.38\%), both chimpanzee and human had at least one read.
+When we restrict the analysis to sites with at least three reads, 109,098 sites (93.77\%) had sequences in human, 89,628 sites (77.04\%) in chimpanzee, and 86,604 sites (74.44\%) in both.
+From this data, we found 9,275 SNPs relative to the human reference genome that were present in both chimpanzee and human datasets.
 
@@
 		Taxon                  & \# Reads  & \# Filtered Reads & \# Loci $\ge 1$ Read & \# Loci $\ge 3$ Reads & \# SNPs \\ \hline\hline
-		\emph{Pan troglodytes} & 2,616,062 &  &  &  &  \\ \hline
-		\emph{Homo sapiens}    & 3,032,823 &  &  &  &  \\ \hline
+		\emph{Pan troglodytes} & 3,917,046 & 2,826,643           &  91,646              &  89,628               & 309,703 \\ \hline
+		\emph{Homo sapiens}    & 4,542,978 & 3,784,192           & 111,686              & 109,098               &  35,651 \\ \hline
 	\end{tabular}

### 2012-12-14 paper/paper.tex
Added abstract and discussion from Luca

@@
 \begin{abstract}
-\textcolor{red}{\ldots (This is just a bit of the blurb from the email.) Our paper is an introduction to a 2nd generation sequencing technique for typing thousands of genome-wide markers from non-model organisms.
-We demonstrate it in five primates and discuss its promise for doing mutli-locus phylogenetics and population genetics in primates.
-\ldots}
+Over the last two decades primatologists have benefited from the use of numerous molecular markers to study various aspects of primate behavior and evolutionary history.
+However, most of the studies to date have been based on a single locus (usually mitochondrial DNA) or a few nuclear markers (e.g., microsatellites).
+Unfortunately, the use of such markers is not only unable to successfully address important questions in primate population genetics and phylogenetics (mainly due to the discordance between gene tree and species tree), but their development is often a time-consuming and expensive task.
+Today, the advent of next-generation sequencing (NGS) allows researchers to generate large amount of genomic data for non model organisms.
+However, whole genome sequencing is still cost prohibitive for most primate species.
+In this paper, we introduce a second generation sequencing technique for typing thousands of genome-wide markers for non-model organisms.
+Restriction site-Associated DNA Sequencing (RAD-Seq) reduces the complexity of the genome and allow cheap and fast discovery of thousands of markers in many individuals.
+Here, we describe the principles of this technique and we demonstrate its application in five primates, representing some of the major lineages within the order.
+Finally, we discuss the promise and possible limitations of such method for doing multi-locus phylogenetics and population genetics in primates.
 \end{abstract}
@@
 
-[What is mol.
-primatology?
-Explain importance of developing markers across taxa in order to study primates at different levels (population, phylogeography, phylogenetics)] During the first decades of molecular primatology, an impediment to researchers was the need to develop and type polymorphic markers in a taxon of interest.
+In molecular primatology, as in all fields of molecular biology, the development of variable genetic markers is essential for the study of organisms at different levels, from population genetic to phylogeographic to phylogenetic research.
+During the first decades of the discipline, an impediment to researchers was the need to develop and type polymorphic markers in a taxon of interest.
 The markers that resulted from this time-consuming and expensive task were often uninformative when applied outside of the population used in their design, necessitating further rounds of primer design or microsatellite assays, for example (Davey et al., 2011).
-\textcolor{red}{Due to the bottleneck of marker discovery}, many population genetic or phylogenetic studies in molecular primatology have been based on one or few loci, usually mitochondrial DNA or few microsatellites (\textcolor{red}{Ting and Sterner, 2012}).
+Due to the bottleneck caused by the inefficiency of marker discovery, many population genetic or phylogenetic studies in molecular primatology have been based on one or few loci, usually mitochondrial DNA or a few microsatellites (Ting and Sterner, 2012).
 Such inferences can reliably give the evolutionary history of those particular regions of the genome, but they fail to adequately capture the complete complex history of the population given the mosaic nature of genomic evolution (Degnan and Rosenberg, 2009; Maddison, 1997; Edwards, 2009; Maddison and Knowles, 2006).
@@
 
-The following is a summary of the RAD tag library preparation of Etter et al., 2011 (\textcolor{red}{Fig.
-1}).
-The RAD-seq library preparation begins when genomic DNA from multiple individuals is digested with a restriction enzyme, such as \emph{EcoRI} or \emph{PspXI} (1a).
+The following is a summary of the RAD tag library preparation protocol of Etter et al., 2011 (Fig.
+1).
+The RAD-seq library preparation begins when genomic DNA is digested with a restriction enzyme, such as \emph{EcoRI} or \emph{PspXI} (1A).
 The P1 adapter is then ligated to the fragments, connected to the sticky end at the restriction enzyme cut site.
-The P1 adapter contains an amplification site for PCR, an Illumina sequencing priming site, and an individual-specific barcode of five basepairs (1b).
-Once the barcode has been added, fragments from multiple individuals can be pooled, and the DNA is randomly sheared with a sonicator to have a length distribution under 1 kilobase (1c).
+The P1 adapter contains an amplification site for PCR, an Illumina sequencing priming site, and an individual-specific barcode of five basepairs (1B).
+Once the barcode has been added, fragments from multiple individuals can be pooled (1C), and the DNA is randomly sheared with a sonicator to have a length distribution under 1 kilobase (1D).
 To select for reads that are suitable for sequencing on the Illumina platform, the sheared samples are size selected via agarose gel electrophoresis, extracting fragments between 300 and 500 bp in length.
-The second adapter, P2, is a Y adapter meaning its two halves are complementary for only part of their length (1d).
-It is ligated to the fragments and then the fragments are amplified via PCR (1e).
+The second adapter, P2, is a Y adapter meaning its two halves are complementary for only part of their length (1E).
+It is ligated to the fragments and then the fragments are amplified via PCR (1F).
 Because the second adapter has divergent ends, the reverse amplification primer is unable to bind until after the forward amplification primer has filled in its complementary sequence.
@@
 
+As input for the clustering analysis, we combined the first read of the paired-end run and the single-end run reads.
 We demultiplexed, or separated by barcode, sequence reads and excluded from the analysis reads without an expected barcode or an intact restriction enzyme cut site.
 We also removed reads with any quality scores below 10.
-Using the program Stacks, we clustered all reads into matching sets and compared closely related sets to detect orthologous loci and SNPs using a maximum likelihood approach (\textcolor{red}{Stacks ref}).
-We tallied orthologous SNPs using VCFtools (\textcolor{red}{ref}).
-
-\textcolor{red}{[Phylogeny?]}
+Using the program Stacks, we clustered all reads into sets that differed by no more than two basepairs and compared closely related sets to detect orthologous loci and SNPs using a maximum likelihood approach (Catchen et al., 2011).
+We tallied orthologous SNPs using VCFtools (Danecek et al., 2011).
 
@@
 
-To assess the RAD tag coverage, we mapped human and chimpanzee reads to the human reference genome.
+To assess the RAD tag coverage, we mapped human and chimpanzee reads to the highest quality primate reference genome, that of humans.
 Again, we excluded reads without an expected barcode or an intact restriction enzyme cut site.
-We aligned reads to the human reference genome (\textcolor{red}{hg19, ref}) using BWA with default parameters.
-We removed reads that were unmapped or that had low mapping quality after alignment using Picard (\textcolor{red}{ref}) and BamTools (\textcolor{red}{ref}).
+We aligned reads to the human reference genome (GRCh37/hg19, Human Genome Consortium 2001) using BWA with default parameters (Li and Durbin, 2009).
+We separately mapped the single-end and paired-end data and then combined the resultant files after alignment.
+We removed reads that were unmapped or that had low mapping quality using Picard (http://picard.sourceforge.net) and BamTools (Barnett et al., 2011).
 
-\textcolor{red}{Cut this paragraph?} After performing local realignment around indels with GATK (\textcolor{red}{ref}), we identified SNPs and short indels using SAMtools mpileup and BCFtools (\textcolor{red}{ref}).
+After performing local realignment around indels with GATK (DePristo et al., 2011), we identified SNPs and short indels using SAMtools mpileup and BCFtools (Li et al., 2009).
 We required a minimum coverage of 3 reads and a maximum of 100 to call a SNP or an indel at a given location.
-We tallied orthologous SNPs using VCFtools (\textcolor{red}{ref}).
+We tallied orthologous SNPs using VCFtools (Danecek et al., 2011).
 
-To assess how many restriction sites were successfully sequenced and to analyze the degree of overlap between multiplexed individual's datasets, we first found all possible \emph{PspXI} cut sites in the human genome using the oligoMatch utility in the USCS Genome Browser program (\textcolor{red}{ref}).
-This allowed us to calculate the coverage of these restriction site-associated regions using BEDtools' multiBamCov program (\textcolor{red}{ref}).
+To assess how many restriction sites were successfully sequenced and to analyze the degree of overlap between multiplexed individual's datasets, we first found all possible \emph{PspXI} cut sites in the human genome using the oligoMatch utility in the USCS Genome Browser program and created a BED file of all regions 1000 basepairs upstream and downstream.
+(Meyer et al., 2012).
+This allowed us to calculate the coverage of these restriction site-associated regions using BEDtools' multiBamCov program (Quinlan and Hall 2010).
 
@@
 
-% 3804227030 bp total (12,983,710 pairs of reads * (142bp + 151bp). This includes non-OWM taxa, remember.
-% OWM sequencing reads: 6,125,860 total, half that many pairs
-% Of those, 2029586 passed the filtration steps
-
-\_\_\_ million sequencing reads with an intact barcode and restriction enzyme cut site could be assigned confidently to one of the five primates in the present study.
-Roughly \_\_\_ million of those reads passed quality control filtration and \_\_\_ could be assigned to a stack.
-In these stacks our study identified \_\_\_ SNPs among all samples.
+12.3 million sequencing reads with an intact barcode and restriction enzyme cut site could be assigned confidently to one of the five primates in the present study.
+Roughly 9.1 million of those reads passed quality control filtration and were clustered into stacks.
+By comparing these stacks and including only SNPs that were present in multiple individuals, our study identified 7,910 SNPs among all samples.
 Information for each individual is summarized in Table 1.
@@
 		\hline
-		Taxon & Source & \# Reads & \# Filtered Reads & \# Stacks & Mean Coverage & \# SNPs \\ \hline\hline
-		\emph{Microcebus \textcolor{red}{murinus?}} & \textcolor{red}{Unknown} &  &  &  &  &  \\ \hline
-		\emph{Cebus \textcolor{red}{sp.?}} & \textcolor{red}{Unknown} &  &  &  &  &  \\ \hline
-		\emph{Theropithecus gelada} & \textcolor{red}{Unknown} &  &  &  &  &  \\ \hline
-		\emph{Pan troglodytes} & \textcolor{red}{Unknown} &  &  &  &  &  \\ \hline
-		\emph{Homo sapiens} & \textcolor{red}{Unknown} &  &  &  &  &  \\ \hline
-		\hline
+		Taxon                       & \# Reads  & \# Filtered Reads & \# Ind. Stacks & Mean Coverage (SD)  & \# Shared SNPs \\ \hline\hline
+		\emph{Microcebus sp.}       & 2,830,832 & 2,025,103       & 248,324         &  7.66 (20.71)  & 13    \\ \hline
+		\emph{Cebus sp.}            & 1,946,096 & 1,427,413       & 107,829         & 12.64 (139.08) & 56    \\ \hline
+		\emph{Theropithecus gelada} & 1,918,425 & 1,392,709       & 136,657         &  9.70 (17.34)  & 212   \\ \hline
+		\emph{Pan troglodytes}      & 2,616,062 & 1,910,560       & 157,775         & 11.50 (42.27)  & 5,886 \\ \hline
+		\emph{Homo sapiens}         & 3,032,823 & 2,374,733       & 131,544         & 17.37 (25.30)  & 5,786 \\ \hline
 	\end{tabular}
@@
 
-In the human genome, we found \_\_\_ possible cut sites for \emph{PspXI} and \_\_\_ possible sequencing sites (two per cut site, one upstream and one downstream).
-Of those, \_\_\_ sites (\_\_\_\%) were covered by at least one read in human and \_\_\_ sites (\%) in chimpanzee.
-For \_\_\_ sites (\_\_\%), both chimpanzee and human had at least one read.
-When we restrict the analysis to sites with at least three reads, \_\_\_ sites (\_\_\_\%) were covered in human, \_\_\_ sites (\_\_\_\%) were covered in chimpanzee, and \_\_\_ sites (\_\_\_\%) in both.
+In the human genome, we found 58,172 possible cut sites for \emph{PspXI} and 116,344 possible sequencing sites (two per cut site, one upstream and one downstream).
+Of those possible location, 111,686 sites (96.00\%) had at least one mapped read present in human and 91,646 sites (78.77\%) in chimpanzee.
+For 90,022 sites (77.38\%), both chimpanzee and human had at least one read.
+When we restrict the analysis to sites with at least three reads, 109,098 sites (93.77\%) had sequences in human, 89,628 sites (77.04\%) in chimpanzee, and 86,604 sites (74.44\%) in both.
+From this data, we found 9,275 SNPs relative to the human reference genome that were present in both chimpanzee and human datasets.
 
@@
 		Taxon                  & \# Reads  & \# Filtered Reads & \# Loci $\ge 1$ Read & \# Loci $\ge 3$ Reads & \# SNPs \\ \hline\hline
-		\emph{Pan troglodytes} &  &  &  &  &  \\ \hline
-		\emph{Homo sapiens} &  &  &  &  &  \\ \hline
-		\hline
+		\emph{Pan troglodytes} & 3,917,046 & 2,826,643           &  91,646              &  89,628               & 309,703 \\ \hline
+		\emph{Homo sapiens}    & 4,542,978 & 3,784,192           & 111,686              & 109,098               &  35,651 \\ \hline
 	\end{tabular}
@@
 
-%\begin{itemize}
-%	\item Pairwise comparisons between individuals? Like X orthologous regions shared between Papio and Thero?
-%\end{itemize}
-%
-%%\includegraphics{figs/seq_site_coverage_by_ind}
-%
-%\begin{itemize}
-%	\item SNP info from merged analysis
-%	\item SNP Venn diagram?
-%	\item Count orthologous SNPs shared between individuals. Pairwise?
-%\end{itemize}
-
 \section{Discussion}
 
-\textcolor{red}{[Why it would be good in primates.]}
+RAD-Seq represents a fast, cheap, and efficient method for SNP discovery also in species for which no genome reference is available.
+These characteristics makes RAD-seq an extremely promising techniques for researchers interested in primate population genomics and phylogentics.
+There are several advantages in using RAD-seq over other molecular techniques.
+First, this methodology is quite cheap and requires little labwork.
+The development of a library can be completed in only two days of labwork and all the different steps can be easily performed in a standard molecular lab.
+Also, the possibility to multiplex several individuals in the same Illumina run either using standard barcodes or a custom combinatorial indexing method (see Peterson et al., 2012 for details) allow researchers to reduce the number of sequencing runs decreasing the costs even further (Peterson et al., 2012; Davey et al., 2011; McCormack, et al.
+2012).
+Second, RAD-seq represents a great improvement in discovering molecular markers to be used in population genetics and phylogenetics.
+Previous studies to date have been based on single locus (mainly mitochondrial DNA) or a few tens of loci (microsatellite for population genetics or nuclear loci for phylogenetics).
+RAD-seq techniques can easily produce thousands of independent SNPs in a single run increasing 100-1000x the amount of data available to researchers.
+Finally, RAD sequencing can produce a large amount of orthologous SNP data that can be employed on a wide range of studies, including population genomics and demographics (e.g., effective population size estimates, bottlenecks, etc.), gene flow and hybridization between closely related species, species boundaries, phylogeography, and phylogeny especially at the intrageneric level (Hohenlohe et al., 2010, 2011; Emerson et al., 2010; Keller et al., 2012; Wagner et al., 2012; Rubin et al., 2012).
+In summary, we believe that the use of RAD-seq technology can provide extremely valuable information to study recent radiation within primates and to address some major open questions in primatology.
 
-RADseq is viable tool for researcher interested in primate phylogenetics, pop. gen.
+Despite the great potential of the application of RAD sequencing in primatology, we also need to point out some possible limitations of this technique.
+Possibly, the main constraint of employing RAD-seq on a large scale within primates is related to the need of high quality samples in order to build the library.
+In this study we employed high quality DNA, extracted from tissue or blood.
+However, most molecular primatologists are limited in their use of invasive samples, and more often rely on low quality samples such as hair or feces.
+\textcolor{red}{[I'd like to cut this, since I don't want to hint that we're working on techniques to get RAD tags out of feces] Although not yet available, in theory, RAD-seq protocols using non invasive samples could be developed.
+In a recent study, Perry and colleagues (2010) presented a genomic-scale capture protocol to obtain endogenous DNA from primate fecal samples.
+Capture methods have been also used to obtain low quantity and poor quality DNA from museum specimens (Mason et al., 2011; Guschanski et al., in press) or even fossils (Burbano et al., 2010; Krause et al., 2010).
+It is likely that similar methods can be developed soon allowing primatologists to employ RAD-seq also on low quality DNA samples, such as feces or museum specimens.}
 
-\begin{itemize}
-	\item Cheap and easy to create libraries. Competent lab can do it in two days. (Ha!)
-	\item Multiplexing is great for pop gen studies.
-	\item Enzyme choice allows control over coverage, number of individuals, number of loci.
-	\item Baird: "different marker densities can be attained by choice of restriction enzyme"
-\end{itemize}
+Another possible limitation of the RAD-seq approach is the time scale of its application.
+RAD-seq data in fact might be not suitable for comparing taxa very distantly related (Rubin et al., 2012).
+In their study, Rubin and colleagues showed a negative correlation between phylogenetic accuracy and evolutionary divergence time, supporting age clade as a major determinant of the success of the RAD method (Rubin et al., 2012).
+The deep divergences between taxa in fact decreases the amount of discoverable RAD loci for two main reason: first, restriction sites can change over time, reducing the number of orthologous loci retained across distantly related taxa; second, orthology is more difficult to infer based on sequence similarity when evolutionary divergence is high (Rubin et al., 2012).
+This correlation between accuracy and divergence time either reduces the number of orthologus loci available for phylogenetic reconstruction or increase the amount of missing data; both scenarios can affect phylogenetic performance reducing the support values in many nodes or supporting different topologies.
+However, despite this drawback, Rubin and colleagues were successful in reconstructing the phylogeny of 12 species of \emph{Drosophila}, with a crown age of 40–60 Mya.
+This result suggest that RAD-seq data might be informative enough to reconstruct the phylogeny of most lineages within primates (crown age between 65 and 85Mya – Wilkinson et al., 2011; Perelman et al., 2011; Steiper and Seiffert, 2012).
 
-Changing the restriction enzyme to a more or less frequent cutter allows fine grained control over the number of loci, number of individuals multiplexed, and depth of coverage, allowing researchers to tailor marker discovery to their hypothesis.
-
-Potential problems with RADseq method including the phylogenetic distance and the need of high quality DNA (at least for now)
-
-ddRADseq
+This study illustrates the value of RAD-seq approach in discovering a large number of independent SNPs that can be used to address several questions in primatology, ranging from population genomics to phylogenetics.
+Our preliminary study on primates show the feasibility of this techniques across the primate order, even when nearby reference genomes are not available.
+Further developments in both sequencing technologies and computational tools will address - and most likely overcome - the current limitations of RAD sequencing, making this technique viable for studies on a large number of primate species and populations.
 
 \section{Acknowledgements}
-\begin{itemize}
-	\item NYU Med School folk
-	\item \textcolor{red}{Grants?}
-\end{itemize}
+
+\textcolor{red}{Leakey Foundation Grant} The authors would like to thank the NYU Langone Medical Center's Genome Technology Center for assistance with library preparation and sequencing.
 

### 2012-12-14 paper/paper.tex
Done except for Andy's edits

@@
 
-In molecular primatology, as in all fields of molecular biology, the development of variable genetic markers is essential for the study of organisms at different levels, from population genetic to phylogeographic to phylogenetic research.
+In molecular primatology, as in all fields of molecular biology, the development of variable genetic markers is essential for the study of organisms at different levels, from population genetic to phylogeographic to phylogenetic research (Avise, 1994).
 During the first decades of the discipline, an impediment to researchers was the need to develop and type polymorphic markers in a taxon of interest.
@@
 	We digested genomic DNA from 5 primates with \emph{PspXI} (New England Biolabs) and used it to create a multiplexed RAD tag library.
-Our library preparation method followed that of Etter et al, 2011 with the following modifications: the P1 adapter top\textcolor{red}{(?)} oligonucleotide was modified to have an overhang corresponding to the cut site of \emph{PspXI}, and a longer P2 adapter suitable for paired end sequencing was used (P2\_top: 5'-\textcolor{red}{SEQUENCEHERE}-3'; P2\_bottom: 5'-\textcolor{red}{SEQUENCEHERE}-3').
-Individual-specific barcodes contained in the P1 adapter differed by at least \textcolor{red}{three} nucleotides.
+Our library preparation method followed that of Etter et al, 2011 with the following modifications: the P1 adapter bottom oligonucleotide was modified to have an overhang corresponding to the cut site of \emph{PspXI}, and a longer P2 adapter suitable for paired end sequencing was used (P2\_top: 5'- /5Phos/GAT CGG AAG AGC GGT TCA GCA GGA ATG CCG AGA CCG ATC AGA ACA A-3'; P2\_bottom: 5'- CAA GCA GAA GAC GGC ATA CGA GAT CGG TCT CGG CAT TCC TGC TGA ACC GCT CTT CCG ATC*T -3').
+Individual-specific barcodes contained in the P1 adapter differed by at least three nucleotides.
 We chose \emph{PspXI} based on the results of \emph{in silico} digestion of the human, rhesus macaque, and baboon reference genomes using custom Perl scripts.
@@
 Other individuals were sequenced alongside those of the present study.
-Sequences are available to download from the NCBI Short Read Archive (accession number \textcolor{red}{SRAXXXXXX.X}).
+Sequences are available to download from the NCBI Short Read Archive (accession number SRAXXXXXX.X).
 
@@
 However, most molecular primatologists are limited in their use of invasive samples, and more often rely on low quality samples such as hair or feces.
-\textcolor{red}{[I'd like to cut this, since I don't want to hint that we're working on techniques to get RAD tags out of feces] Although not yet available, in theory, RAD-seq protocols using non invasive samples could be developed.
+Although not yet available, in theory, RAD-seq protocols using non invasive samples could be developed.
 In a recent study, Perry and colleagues (2010) presented a genomic-scale capture protocol to obtain endogenous DNA from primate fecal samples.
 Capture methods have been also used to obtain low quantity and poor quality DNA from museum specimens (Mason et al., 2011; Guschanski et al., in press) or even fossils (Burbano et al., 2010; Krause et al., 2010).
-It is likely that similar methods can be developed soon allowing primatologists to employ RAD-seq also on low quality DNA samples, such as feces or museum specimens.}
 
 Another possible limitation of the RAD-seq approach is the time scale of its application.
-RAD-seq data in fact might be not suitable for comparing taxa very distantly related (Rubin et al., 2012).
-In their study, Rubin and colleagues showed a negative correlation between phylogenetic accuracy and evolutionary divergence time, supporting age clade as a major determinant of the success of the RAD method (Rubin et al., 2012).
+RAD-seq data in fact might be not suitable for comparing very distantly related taxa (Rubin et al., 2012).
+In their study, Rubin and colleagues showed a negative correlation between phylogenetic accuracy and evolutionary divergence time, supporting age clade as a critical factor in the success of the RAD method (Rubin et al., 2012).
 The deep divergences between taxa in fact decreases the amount of discoverable RAD loci for two main reason: first, restriction sites can change over time, reducing the number of orthologous loci retained across distantly related taxa; second, orthology is more difficult to infer based on sequence similarity when evolutionary divergence is high (Rubin et al., 2012).
@@
 However, despite this drawback, Rubin and colleagues were successful in reconstructing the phylogeny of 12 species of \emph{Drosophila}, with a crown age of 40–60 Mya.
-This result suggest that RAD-seq data might be informative enough to reconstruct the phylogeny of most lineages within primates (crown age between 65 and 85Mya – Wilkinson et al., 2011; Perelman et al., 2011; Steiper and Seiffert, 2012).
+This result suggest that RAD-seq data might be informative enough to reconstruct the phylogeny of most lineages within primates (crown age between 65 and 85 Mya – Wilkinson et al., 2011; Perelman et al., 2011; Steiper and Seiffert, 2012).
 
@@
 Our preliminary study on primates show the feasibility of this techniques across the primate order, even when nearby reference genomes are not available.
-Further developments in both sequencing technologies and computational tools will address - and most likely overcome - the current limitations of RAD sequencing, making this technique viable for studies on a large number of primate species and populations.
+Future developments in both sequencing technologies and computational tools will address - and most likely overcome - the current limitations of RAD sequencing, making this technique viable for studies on a large number of primate species and populations.
 
@@
 
-\textcolor{red}{Leakey Foundation Grant} The authors would like to thank the NYU Langone Medical Center's Genome Technology Center for assistance with library preparation and sequencing.
+The present study was supported by a Leakey Foundation General Grant.
+The authors would like to thank the NYU Langone Medical Center's Genome Technology Center for assistance with library preparation and sequencing.
 

### 2012-12-14 paper/paper.tex
Finished version, converted to .doc and submitted to IJP

@@
-\title{A new method for genome-wide marker development and typing holds great promise for primate population genetics}
+\title{A new method for genome-wide marker development and typing holds great promise for molecular primatology}
 \author{
@@
 RAD-seq is one such simple, inexpensive reduced representation technique which allows for the sequencing of small fragments of the genome adjacent to restriction enzyme cut sites (Baird et al., 2008).
-These RAD tags, or restriction-site association DNA tags, were originally developed for use in microarray hybridization typing (Miller et al., 2007), but an updated protocol substitutes second-generation DNA sequencing to rapidly discover and type SNPs (Baird et al., 2008; Etter et al., 2011).
+These RAD tags, or Restriction-site Association DNA tags, were originally developed for use in microarray hybridization typing (Miller et al., 2007), but an updated protocol substitutes second-generation DNA sequencing to rapidly discover and type SNPs (Baird et al., 2008; Etter et al., 2011).
 The lack of reliance on a reference genome and applicability to datasets of many individuals make it a promising technique for phylogenetic or population genetic studies in non-model organisms, such as many primates.
@@
 This ensures that only RAD tags ligated to P1 are able to amplify.
-After few cycles of PCR to minimize the risk of introducing PCR artifacts or biases, the library is ready for final clean-up, quality control, and sequencing.
+After few a cycles of PCR to minimize the risk of introducing PCR artifacts or biases, the library is ready for final clean-up, quality control, and sequencing.
 
@@
 
-Restriction site-associated DNA Sequencing (RAD-Seq) is an economical and efficient method for SNP discovery and genotyping.
+RAD-Seq is an economical and efficient method for SNP discovery and genotyping.
 Since its first application by Baird and colleagues (2008) on two model organisms – the fungus \emph{Neurospora crassa} and the three-spined stickleback, \emph{Gasterosteus aculeatus} – RAD-seq has been successfully applied to several organisms for which reference genome information was not available.
@@
 For example, a series of studies conducted by Hohenlohe and colleagues investigated parallel adaptation and hybridization in several species of fish (Hohenlohe et al.
-2010, 2011, 2012), while Emerson et al (2010) identified more than 3,700 SNPs for pitcher plant mosquitoes in eastern North America to provide the first phylogeographic study using RAD sequence data.
+2010, 2011, 2012), while Emerson et al (2010) identified more than 3,700 SNPs for pitcher plant mosquitoes in eastern North America, providing the first phylogeographic study using RAD sequence data.
 
@@
 Rubin et al (2012) provided a simulation study in which they investigated the accuracy of RAD-seq data to reconstruct phylogenies in organisms with different population sizes and clade ages (\emph{Drosophila}, mammals, and yeasts).
-In their study the authors supported the efficiency of RAD-seq data in inferring phylogenies, but they also caution about how this approach achieves the best results in younger clades where more orthologous restriction sites are likely to be retained across species.
-This simulation analysis was confirmed in two recent empirical studies in which a RAD-tag sequencing approach was successfully used to reconstruct phylogenetic relationships in two recent but speciose radiation, such as the African cyclids (Wagner et al., 2012) and the \emph{Heliconius} butterﬂies (Nadeau et al., 2012).
+In their study the authors supported the efficiency of RAD-seq data in inferring phylogenies, but they also caution that this approach achieves the best results in younger clades where more orthologous restriction sites are likely to be retained across species.
+This simulation analysis was confirmed in two recent empirical studies in which a RAD-tag sequencing approach was successfully used to reconstruct phylogenetic relationships in two recent but speciose radiations: the African cyclids (Wagner et al., 2012) and the \emph{Heliconius} butterﬂies (Nadeau et al., 2012).
 
@@
 
-To assess how many restriction sites were successfully sequenced and to analyze the degree of overlap between multiplexed individual's datasets, we first found all possible \emph{PspXI} cut sites in the human genome using the oligoMatch utility in the USCS Genome Browser program and created a BED file of all regions 1000 basepairs upstream and downstream.
+To assess how many restriction sites were successfully sequenced and to analyze the degree of overlap between multiplexed individuals' datasets, we first found all possible \emph{PspXI} cut sites in the human genome using the oligoMatch utility in the USCS Genome Browser program and created a BED file of all regions 1000 basepairs upstream and downstream.
 (Meyer et al., 2012).
@@
 When we restrict the analysis to sites with at least three reads, 109,098 sites (93.77\%) had sequences in human, 89,628 sites (77.04\%) in chimpanzee, and 86,604 sites (74.44\%) in both.
-From this data, we found 9,275 SNPs relative to the human reference genome that were present in both chimpanzee and human datasets.
+From these data, we found 9,275 SNPs relative to the human reference genome that were present in both chimpanzee and human datasets.
 
@@
 
-RAD-Seq represents a fast, cheap, and efficient method for SNP discovery also in species for which no genome reference is available.
-These characteristics makes RAD-seq an extremely promising techniques for researchers interested in primate population genomics and phylogentics.
+RAD-Seq represents a fast, cheap, and efficient method for SNP discovery even in species for which no genome reference is available.
+These characteristics make RAD-seq an extremely promising technique for researchers interested in primate population genomics and phylogenetics.
 There are several advantages in using RAD-seq over other molecular techniques.
@@
 The development of a library can be completed in only two days of labwork and all the different steps can be easily performed in a standard molecular lab.
-Also, the possibility to multiplex several individuals in the same Illumina run either using standard barcodes or a custom combinatorial indexing method (see Peterson et al., 2012 for details) allow researchers to reduce the number of sequencing runs decreasing the costs even further (Peterson et al., 2012; Davey et al., 2011; McCormack, et al.
+Also, the possibility to multiplex several individuals in the same Illumina run either using standard barcodes or a custom combinatorial indexing method (see Peterson et al., 2012 for details) allows researchers to reduce the number of sequencing runs, decreasing the costs even further (Peterson et al., 2012; Davey et al., 2011; McCormack, et al.
 2012).
-Second, RAD-seq represents a great improvement in discovering molecular markers to be used in population genetics and phylogenetics.
+Second, RAD-seq represents a great improvement in the discovery of molecular markers to be used in population genetics and phylogenetics.
 Previous studies to date have been based on single locus (mainly mitochondrial DNA) or a few tens of loci (microsatellite for population genetics or nuclear loci for phylogenetics).
 RAD-seq techniques can easily produce thousands of independent SNPs in a single run increasing 100-1000x the amount of data available to researchers.
-Finally, RAD sequencing can produce a large amount of orthologous SNP data that can be employed on a wide range of studies, including population genomics and demographics (e.g., effective population size estimates, bottlenecks, etc.), gene flow and hybridization between closely related species, species boundaries, phylogeography, and phylogeny especially at the intrageneric level (Hohenlohe et al., 2010, 2011; Emerson et al., 2010; Keller et al., 2012; Wagner et al., 2012; Rubin et al., 2012).
-In summary, we believe that the use of RAD-seq technology can provide extremely valuable information to study recent radiation within primates and to address some major open questions in primatology.
+Finally, RAD sequencing can produce a large amount of orthologous SNP data that can be employed in a wide range of studies, including population genomics and demographics (e.g., effective population size estimates, bottlenecks, etc.), gene flow and hybridization between closely related species, species boundaries, phylogeography, and phylogeny especially at the intrageneric level (Hohenlohe et al., 2010, 2011; Emerson et al., 2010; Keller et al., 2012; Wagner et al., 2012; Rubin et al., 2012).
+In summary, we believe that the use of RAD-seq technology will provide extremely valuable information to study recent radiation within primates and to address some major open questions in primatology.
 
 Despite the great potential of the application of RAD sequencing in primatology, we also need to point out some possible limitations of this technique.
-Possibly, the main constraint of employing RAD-seq on a large scale within primates is related to the need of high quality samples in order to build the library.
-In this study we employed high quality DNA, extracted from tissue or blood.
+Possibly, the main constraint of employing RAD-seq on a large scale within primates is related to the need for high quality DNA in order to build the library.
+In this study we used DNA extracted from tissue or blood.
 However, most molecular primatologists are limited in their use of invasive samples, and more often rely on low quality samples such as hair or feces.
@@
 In a recent study, Perry and colleagues (2010) presented a genomic-scale capture protocol to obtain endogenous DNA from primate fecal samples.
-Capture methods have been also used to obtain low quantity and poor quality DNA from museum specimens (Mason et al., 2011; Guschanski et al., in press) or even fossils (Burbano et al., 2010; Krause et al., 2010).
+Capture methods have been also used to obtain low quantity and poor quality DNA from museum specimens (Mason et al., 2011) or even fossils (Burbano et al., 2010; Krause et al., 2010).
 
-Another possible limitation of the RAD-seq approach is the time scale of its application.
+Another possible limitation of the RAD-seq approach is the evolutionary time scale of its application.
 RAD-seq data in fact might be not suitable for comparing very distantly related taxa (Rubin et al., 2012).
-In their study, Rubin and colleagues showed a negative correlation between phylogenetic accuracy and evolutionary divergence time, supporting age clade as a critical factor in the success of the RAD method (Rubin et al., 2012).
-The deep divergences between taxa in fact decreases the amount of discoverable RAD loci for two main reason: first, restriction sites can change over time, reducing the number of orthologous loci retained across distantly related taxa; second, orthology is more difficult to infer based on sequence similarity when evolutionary divergence is high (Rubin et al., 2012).
+In their study, Rubin and colleagues showed a negative correlation between phylogenetic accuracy and evolutionary divergence time, suggesting that the age of a clade is a major determinant of the success of the RAD method (Rubin et al., 2012).
+The deep divergences between taxa in fact decreases the number of discoverable RAD loci for two main reasons: first, restriction sites can change over time, reducing the number of orthologous loci retained across distantly related taxa; second, orthology is more difficult to infer based on sequence similarity when evolutionary divergence is high (Rubin et al., 2012).
 This correlation between accuracy and divergence time either reduces the number of orthologus loci available for phylogenetic reconstruction or increase the amount of missing data; both scenarios can affect phylogenetic performance reducing the support values in many nodes or supporting different topologies.
-However, despite this drawback, Rubin and colleagues were successful in reconstructing the phylogeny of 12 species of \emph{Drosophila}, with a crown age of 40–60 Mya.
-This result suggest that RAD-seq data might be informative enough to reconstruct the phylogeny of most lineages within primates (crown age between 65 and 85 Mya – Wilkinson et al., 2011; Perelman et al., 2011; Steiper and Seiffert, 2012).
+However, despite this drawback, Rubin and colleagues successfully reconstructed the phylogeny of 12 species of \emph{Drosophila}, with a crown age of 40–60 Mya.
+This result suggests that RAD-seq data might be informative enough to reconstruct the phylogeny of most lineages within primates (crown age between 65 and 85 Mya – Wilkinson et al., 2011; Perelman et al., 2011; Steiper and Seiffert, 2012).
 
-This study illustrates the value of RAD-seq approach in discovering a large number of independent SNPs that can be used to address several questions in primatology, ranging from population genomics to phylogenetics.
-Our preliminary study on primates show the feasibility of this techniques across the primate order, even when nearby reference genomes are not available.
-Future developments in both sequencing technologies and computational tools will address - and most likely overcome - the current limitations of RAD sequencing, making this technique viable for studies on a large number of primate species and populations.
+This study illustrates the value of RAD-seq approach in discovering a large number of independent SNPs that can be used to address many questions in primatology, ranging from population genomics to phylogenetics.
+Our preliminary study of primates shows the feasibility of this technique across the primate order, even when nearby reference genomes are not available.
+Future developments in both sequencing technologies and computational tools will address - and most likely overcome - the current limitations of RAD sequencing, making this technique viable for studies of a large number of primate species and populations.
 
@@
 
+\section{Figure Legends}
+
+\paragraph{Figure 1:} An overview of the RAD-seq library creation protocol and initial analysis steps.
+
 \end{document}

### 2012-12-14 trunk/scripts/analyze_enzyme_count_data.R
Code used to submit IJP paper.

@@ -1,8 +1,8 @@
 
 rad_counts=read.table("RAD_coverage.txt")
-names(rad_counts)=c("chr","start","end", "enz", "len", "str", "Allenopithecus", "Coast_langur", "CTA", "Mmulatta1", "Papio19349", "Tgelada871162")
+names(rad_counts)=c("chr","start","end", "enz", "len", "str", "H_sapiens", "P_trog")
 
-rad_counts_only = rad_counts[,7:12]
+rad_counts_only = rad_counts[,7:8]
 
 rad_present = rad_counts_only > 0
 rad_gt3     = rad_counts_only >= 3
@@ -10,23 +10,7 @@ rad_gt3     = rad_counts_only >= 3
 rad_sums = rowSums(rad_present)
 rad_sums_gt3 = rowSums(rad_gt3)
 
-#	length(rad_sums[rad_sums == 6]) # 13456
-#	length(rad_sums[rad_sums == 5]) # 23350
-#	length(rad_sums[rad_sums == 4]) # 22890
-#	length(rad_sums[rad_sums == 3]) # 18488
-#	length(rad_sums[rad_sums == 2]) # 12136
-#	length(rad_sums[rad_sums == 1]) # 10818
-#	length(rad_sums[rad_sums == 0]) # 7590
-#	
-#	length(rad_sums[rad_sums >= 6]) # 13456
-#	length(rad_sums[rad_sums >= 5]) # 36806
-#	length(rad_sums[rad_sums >= 4]) # 59696
-#	length(rad_sums[rad_sums >= 3]) # 78184
-#	length(rad_sums[rad_sums >= 2]) # 90320
-#	length(rad_sums[rad_sums >= 1]) # 101138
-#	length(rad_sums[rad_sums >= 0]) # 108728
-
-x = 0:6
+x = 0:2
 
 ### One or more reads --------------------------------------------------------------------
 

### 2012-12-14 trunk/scripts/convert_fasta_to_phylip.pl

@@ -10,9 +10,9 @@ use lib '/home/cmb433/local_perl/';
 
 use Bio::AlignIO;
 
-my $in  = Bio::AlignIO->new(	-file   => "results/merged.align.fasta",
+my $in  = Bio::AlignIO->new(	-file   => "results/stacks/batch_1.align.fasta",
 								-format => "fasta");
-my $out = Bio::AlignIO->new(	-file => ">results/merged.align.phy",
+my $out = Bio::AlignIO->new(	-file => ">results/stacks/batch_1.align.phy",
 								-format => "phylip");
  
 while (my $aln = $in->next_aln) { 

### 2012-12-14 trunk/scripts/merge_vcf.sh

@@ -6,7 +6,7 @@
 
 export PATH=$PATH:$TABIX
 
-ALL_VCF_GZ=$(ls results/*.bwa.rhesus.passed.realn.bam.RG.flt.vcf.gz)
+ALL_VCF_GZ=$(ls results/*.bwa.${GENOME_NAME}.passed.realn.flt.vcf.gz)
 
 ${VCFTOOLS}/vcf-merge \
 	${ALL_VCF_GZ} \

### 2012-12-14 trunk/scripts/vcf_tab_to_fasta_alignment.pl

@@ -35,7 +35,8 @@ my %iupac = (
 			'./.' => '.',
 		);
 
-my $input_tab = "results/merged.flt.vcf.tab";
+#my $input_tab = "results/merged.flt.vcf.tab";
+my $input_tab = "results/stacks/batch_1.vcf.tab";
 
 open (TAB, "<$input_tab")
 	or die "ERROR: Could not open input file $input_tab.\n";
@@ -53,13 +54,13 @@ LINE: foreach my $line (<TAB>) {
 
 	my @data = split /\t/, $line;
 	
-	# Skip if this is indel (Length of @data will be less than 9)
-	if ((scalar @data) < 9) {
+	# Skip if this is indel (Length of @data will be less than 8)
+	if ((scalar @data) < 8) {
 		next LINE;
 	}
 	
 	# Skip if any basepairs are actually 2 or more together
-	for (my $i = 2; $i < 9; $i++) {
+	for (my $i = 2; $i < 8; $i++) {
 		
 		my $bp = $data[$i]; 
 		chomp $bp;
@@ -68,17 +69,34 @@ LINE: foreach my $line (<TAB>) {
 		}
 	}
 	
-	# Exclude heterozygotes. Keep only fixed SNPs
-	for (my $i = 2; $i < 9; $i++) {
-		
-		my $bp = $data[$i]; 
-		chomp $bp;
-		if ($bp =~ /(\w)\/(\w)/) {
-			if ($1 ne $2) {
-				next LINE;
-			}
-		}
-	}
+#		# Exclude heterozygotes. Keep only fixed SNPs
+#		for (my $i = 2; $i < 8; $i++) {
+#			
+#			my $bp = $data[$i]; 
+#			chomp $bp;
+#			if ($bp =~ /(\w)\/(\w)/) {
+#				if ($1 ne $2) {
+#					next LINE;
+#				}
+#			}
+#		}
+#		
+#		# Skip BPs with too much missing data
+#		my $required = 5;
+#	
+#		for (my $i = 2; $i < 8; $i++) {
+#	
+#			my $missing_count = 0;		
+#			my $bp = $data[$i]; 
+#			chomp $bp;
+#			if ($bp eq "./") {
+#				$missing_count++
+#			}
+#			
+#			if (6 - $missing_count < $required) {
+#				next LINE;
+#			}
+#		}
 	
 	# Otherwise write line to pure temporary file
 	print TEMP $line;
[…]

### 2012-12-14 scripts/call_stacks.sh
Code used to submit IJP paper.

@@ -1,0 +1,76 @@
+#!/bin/sh
+
+# ------------------------------------------------------------------------------
+# --- Run Stacks
+# ------------------------------------------------------------------------------
+
+#STACKS=/home/cmb433/exome_macaque/bin/stacks-0.9999/build/bin
+module load stacks/intel/0.9999
+
+mkdir all_primate
+
+# Combine SE and PE data into files in working directory
+
+cat data/female12.read1.fastq      data/female12_SE.fastq      > all_primate/female12.fastq
+cat data/Cebus.read1.fastq         data/Cebus_SE.fastq         > all_primate/Cebus.fastq
+cat data/Tgelada871162.read1.fastq data/Tgelada871162_SE.fastq > all_primate/Tgelada871162.fastq
+cat data/P_trog.read1.fastq        data/P_trog_SE.fastq        > all_primate/P_trog.fastq
+cat data/H_sapiens.read1.fastq     data/H_sapiens_SE.fastq     > all_primate/H_sapiens.fastq
+
+# FASTX - FASTQ Quality Filter
+
+FASTX=/home/cmb433/exome_macaque/bin/fastx
+
+${FASTX}/fastq_quality_filter -v -q 10 -p 100 -Q33 -i all_primate/female12.fastq      -o all_primate/female12.CLEAN.fastq
+${FASTX}/fastq_quality_filter -v -q 10 -p 100 -Q33 -i all_primate/Cebus.fastq         -o all_primate/Cebus.CLEAN.fastq
+${FASTX}/fastq_quality_filter -v -q 10 -p 100 -Q33 -i all_primate/Tgelada871162.fastq -o all_primate/Tgelada871162.CLEAN.fastq
+${FASTX}/fastq_quality_filter -v -q 10 -p 100 -Q33 -i all_primate/P_trog.fastq        -o all_primate/P_trog.CLEAN.fastq
+${FASTX}/fastq_quality_filter -v -q 10 -p 100 -Q33 -i all_primate/H_sapiens.fastq     -o all_primate/H_sapiens.CLEAN.fastq
+
+# Stacks - Make stacks and call SNPs
+# (PBS to do these)
+
+ustacks -t fastq -f all_primate/female12.CLEAN.fastq      -o all_primate/stacks -i 1 -d -r -m 2 -p 12 -M 2
[…]

### 2012-12-14 scripts/convert_fasta_to_nexus.pl

@@ -10,9 +10,9 @@ use lib '/home/cmb433/local_perl/';
 
 use Bio::AlignIO;
 
-my $in  = Bio::AlignIO->new(	-file   => "results/merged.align.fasta",
+my $in  = Bio::AlignIO->new(	-file   => "results/stacks/batch_1.align.fasta",
 								-format => "fasta");
-my $out = Bio::AlignIO->new(	-file => ">results/merged.align.nex",
+my $out = Bio::AlignIO->new(	-file => ">results/stacks/batch_1.align.nexus",
 								-format => "nexus");
  
 while (my $aln = $in->next_aln) { 

### 2012-12-14 scripts/count_restr_enz_reads.sh

@@ -10,5 +10,5 @@ ALL_BAM=$(ls results/*.passed.realn.bam)
 
 ${BEDTOOLS}/multiBamCov \
 	-bams	${ALL_BAM} \
-	-bed data/PspXI_rheMac2.bed \
+	-bed data/PspXI_hg19.bed \
 	> reports/RAD_coverage.txt

### 2012-12-14 scripts/gather_results.R

@@ -1,0 +1,136 @@
+#!/usr/bin/env Rscript 
+
+# ========================================================================================
+# === Program to gather data needed for paper ============================================
+# ========================================================================================
+
+inds = c("female12", "Cebus", "Tgelada871162", "P_trog", "H_sapiens")
+inds_mapped = c("P_trog", "H_sapiens")
+
+# ----------------------------------------------------------------------------------------
+# --- READS: Reads per individual: -------------------------------------------------------
+# ----------------------------------------------------------------------------------------
+
+cat ("# --- READS: Reads per individual: -------------------------------------------------------\n")
+
+total_raw_reads = 0;
+
+for(i in 1:length(inds)) {
+	cmd.read1 = paste ("wc -l data/", inds[i], ".read1.fastq | cut -d' ' -f 1", sep="");
+	reads.read1 = as.numeric(system(cmd.read1, intern=TRUE)) / 4;
+
+	cmd.readSE = paste ("wc -l data/", inds[i], "_SE.fastq | cut -d' ' -f 1", sep="");
+	reads.readSE = as.numeric(system(cmd.readSE, intern=TRUE)) / 4;
+	
+	reads = reads.read1 + reads.readSE
+	
+	# Keep track of all reads from all individuals
+	total_raw_reads = total_raw_reads + reads
+	
+	cat (paste(inds[i], " - individual raw reads:\t", reads, "\n", sep=""));
+}
+
+# ----------------------------------------------------------------------------------------
+# --- READS: Total sequenced reads in all samples ----------------------------------------
+# ----------------------------------------------------------------------------------------
+
+cat ("# --- READS: Total sequenced reads in all samples -----------------------------------------------\n")
+
+cat (paste(total_raw_reads, "\n"))
+
+# ----------------------------------------------------------------------------------------
+# --- STACKS: Filtered Reads per individual ---------------------------------------------
+# ----------------------------------------------------------------------------------------
+
+cat ("# --- STACKS: Filtered Reads per individual: -----------------------------------------------------\n")
+
+total_flt_reads = 0;
+
+for(i in 1:length(inds)) {
+	cmd = paste ("wc -l all_primate/", inds[i], ".CLEAN.fastq | cut -d' ' -f 1", sep="");
+	ind_flt_reads = as.numeric(system(cmd, intern=TRUE)) / 4;
+
[…]

### 2012-12-14 scripts/local_realign_get_targets.sh

@@ -17,6 +17,7 @@ GENOME_FA=$2
 echo "CMD: java -Xmx4g -jar ${GATK}/GenomeAnalysisTK.jar \
 	-T RealignerTargetCreator \
 	--filter_mismatching_base_and_quals \
+	--num_threads 8 \
 	-R ${GENOME_FA} \
 	-o results/${IND_ID}.bwa.${GENOME_CODE}.passed.bam.list \
 	-I results/${IND_ID}.bwa.${GENOME_CODE}.passed.bam";
@@ -24,6 +25,7 @@ echo "CMD: java -Xmx4g -jar ${GATK}/GenomeAnalysisTK.jar \
 java -Xmx4g -jar ${GATK}/GenomeAnalysisTK.jar \
 	-T RealignerTargetCreator \
 	--filter_mismatching_base_and_quals \
+	--num_threads 8 \
 	-R ${GENOME_FA} \
 	-o results/${IND_ID}.bwa.${GENOME_CODE}.passed.bam.list \
 	-I results/${IND_ID}.bwa.${GENOME_CODE}.passed.bam

### 2012-12-14 scripts/count_restr_enz_reads.sh
Cleaned things up a bit. Made shell script for comparative analysis

@@ -4,11 +4,9 @@
 # --- Count sequence coverage (# reads) at restriction enzyme-associated loci
 # ------------------------------------------------------------------------------
 
-BEDTOOLS=/home/cmb433/exome_macaque/bin/BEDTools-Version-2.13.4/bin
-
 ALL_BAM=$(ls results/*.passed.realn.bam)
 
 ${BEDTOOLS}/multiBamCov \
 	-bams	${ALL_BAM} \
-	-bed data/PspXI_hg19.bed \
+	-bed ${RAD_TAG_BED} \
 	> reports/RAD_coverage.txt
